---
title: "Es30"
output: html_document
date: "2026-05-10"
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```


```{r libraries, eval=FALSE}
# Caricamento librerie usate nel corso
library(MASS)       # stepAIC, Boston dataset
library(car)        # vif, diagnostica
library(corrplot)   # visualizzazione matrici di correlazione
library(boot)       # bootstrap
library(ppcor)      # correlazioni parziali
library(GGally) 
library(mvtnorm)
library(ggplot2)
library(skimr)
library(mclust)
```
The data in the boot::urine dataset come from a study of 79 individuals. We are interested in the presence of calcium crystals in urine (absent/present) as a function 
of urine concentration (osmo) and glomerular filtration (cond). Recode categorical variables as factors.

```{r}
data = boot::urine
head(data)
```

```{r}
# remove rows with NA on covariates of interest
urine_clean <- urine[complete.cases(urine[, c("r", "osmo", "cond")]), ]
nrow(urine_clean)
# factorization of the response variable
urine_clean$r = as.factor(urine_clean$r)
levels(urine_clean$r)
```

### 30.1 What are the logits in this study?

The logit in this study is defined as the logarithm of the odds of having calcium 
oxalate crystals present in urine (r = "1") versus absent (r = "0"):

$$
\text{logit}(p_i) = \log\left(\frac{p_i}{1 - p_i}\right) 
= \log\left(\frac{P(r_i = \text{present})}{P(r_i = \text{absent})}\right)
$$

where $p_i$ denotes the probability of presenting crystals for individual $i$. The 
logistic regression model assumes the logit to be a linear function of the available 
covariates — urine osmolarity (`osmo`) and conductivity (`cond`):

$$
\text{logit}(p_i) = \beta_0 + \beta_1 \cdot \text{osmo}_i + \beta_2 \cdot \text{cond}_i
$$

While probabilities $p_i$ are bounded in $[0, 1]$ and odds in $[0, +\infty)$, the 
logit transformation maps them to the entire real line $(-\infty, +\infty)$, making 
linear modeling with potentially unbounded covariates appropriate. The estimated 
coefficients $\hat\beta_1, \hat\beta_2$ describe how the log-odds of crystal presence 
change with a one-unit increase in the corresponding covariate, holding the other constant.


### 29.1 Provide descriptive statistics for the binary indicators to assess the association of each covariate with the response variable of interest (calc). Comment on the results and suggest possible reasons for the presence or absence of an association.

Since both covariates in this study (osmo and cond) are continuous, the descriptive analysis focuses on numerical summaries by response level and boxplots — the analogue of contingency tables for categorical covariates.

### 29.2 For the quantitative variables, produce boxplots and summary statistics by cancer diagnosis.

```{r}
require(gridExtra)
p1 <- ggplot(urine_clean, aes(x = r, y = osmo)) +
geom_boxplot(aes(fill = r), col = "black") +
geom_point(aes(col = r)) +
xlab("Calcium Crystals") + ylab("Urine concentration") +
scale_fill_manual(values=c("#ffe5ab", "#cce8f8")) +
scale_color_manual(values=c("#E69F00", "#56B4E9")) +
theme_bw()
p2 <- ggplot(urine_clean, aes(x = r, y = calc)) +
geom_boxplot(aes(fill = r), col = "black") +
geom_point(aes(col = r)) +
xlab("Calcium Crystals") + ylab("glomerular filtration") +
scale_fill_manual(values=c("#ffe5ab", "#cce8f8")) +
scale_color_manual(values=c("#E69F00", "#56B4E9")) +
theme_bw()
grid.arrange(p1, p2, ncol = 2)
```
