---
title: "exercitation1"
author: "me"
date: "2026-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

# Setup

## Librerie standard

```{r libraries, eval=FALSE}
# Caricamento librerie usate nel corso
library(MASS)       # stepAIC, Boston dataset
library(car)        # vif, diagnostica
library(corrplot)   # visualizzazione matrici di correlazione
library(boot)       # bootstrap
library(ppcor)      # correlazioni parziali
library(GGally) 
library(mvtnorm)
library(ggplot2)
library(skimr)
```

```{r}
head(diamonds)
levels(diamonds$carat)
```


```{r}
labels(diamonds$carat)
```
##descriptiove stats

```{r}
require(skimr)
skim_without_charts(diamonds)
```
Then interpret the values of the median , mean, if the distribution is symmetric, ...


## Do the data come from an observational study or form a randomized experiment? Comment on each output. 

(always respond observational study - i choose some rules to select the sample, and then observe the result. In randomized experiment- i take the sample, divide in two groups: one of them is drugged and i want to compare the two groups)


#Fit a multiple linear regression model to explain gas consumption as a function of the other covariates. First, use the Akaike Information Criterion to select the model. Comment on the results provided by the step function

We perform model selection with the step function with backwards specification. We use the AIC to choose the optimal model. 
```{r}
mod.full = lm(formula = carat ~., data = diamonds)
step(mod.full, direction = "backward")
```
starting point is AIC = ... (explain briefly), in the first step we see that removing the value of width we have an improvement on the AIC, that is a minimal improving of the model, but gives us a model which is more interpretable and simple and fits the data quite good. In the second step we notice that removing any of the other covariates does not improve the AIC. That suggest us that that no other covariate should be removed and the model is fine how it is



##Report and comment on all the results provided by the summary function for the model selected in the previous step
```{r}
mod.sel = lm(formula= carat~cut +price+length+depth, data = diamonds)
summary(mod.sel)
```


The linear regression model explains the carat weight of diamonds using cut, price, length, and depth showing an excellent fit. With an R-squared of 0.971, the model explains about 97% of the variability in carat, and the highly significant F-statistic confirms its overall validity.
The residuals are small and centered around zero, indicating good predictive accuracy, although there is slight asymmetry that may suggest minor outliers. Among the predictors, price, length, and depth all have strong positive and highly significant effects, meaning that larger and more expensive diamonds tend to have greater carat weight. 
The cut variable, included as categorical dummies, has a statistically significant but relatively small negative effect compared to the reference category.
Overall, the model is very effective, with carat mainly driven by price and physical dimensions, while cut plays a secondary role.

##Add the estimated parallel lines to scatterplot of the data and comment on the figure. Write the equations that characterize this model. 
The graphical representation is not done because of the number of dimensions of the regression (it is a plane, not a line). we are just going to write the equation.

first category of cut variable
hat = -1.15 + 0.03*price + 0.16*length + 0.27*depth (cut=1)

second category of cut variable
hat = -1.15 -0.04 + 0.03*price + 0.16*length + 0.27*depth (cut=2)

second category of cut variable
hat = -1.15 -0.05 + 0.03*price + 0.16*length + 0.27*depth (cut=3)

second category of cut variable
hat = -1.15 -0.04 + 0.03*price + 0.16*length + 0.27*depth (cut=4)

second category of cut variable
hat = -1.15 -0.06 + 0.03*price + 0.16*length + 0.27*depth (cut=5)

In practice, all of the coefficient remain the same, the thing that changes is the intercept.


##24.7
```{r}
num_X = data.frame(intercept = 1,cut = as.factor(4), price = 13, length= 6, depth = 4)

predict(mod.sel, num_X, interval = "confidence")
predict(mod.sel, num_X, interval = "prediction")

```
# Confidence interval (per la media condizionale E[Y|X]):
# ŷ(x0) ± t_{n-p, 1-α/2} * σ̂ * sqrt( x0' (X'X)^{-1} x0 )

# Prediction interval (per una nuova osservazione Y_new):
# ŷ(x0) ± t_{n-p, 1-α/2} * σ̂ * sqrt( 1 + x0' (X'X)^{-1} x0 )
