---
title: "Statistical Modelling — Cheat Sheet Esame"
author: "Fra"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
fontsize: 10pt
geometry: margin=2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

# Setup

## Librerie standard

```{r libraries, eval=FALSE}
# Caricamento librerie usate nel corso
library(MASS)       # stepAIC, Boston dataset
library(car)        # vif, diagnostica
library(corrplot)   # visualizzazione matrici di correlazione
library(boot)       # bootstrap
library(ppcor)      # correlazioni parziali
library(GGally) 
library(mvtnorm)
library(ggplot2)
library(skimr)
library(gridExtra)

```

## Import dati tipico

```{r import-template, eval=FALSE}
# Da CSV
df <- read.csv("path/to/file.csv", header = TRUE, stringsAsFactors = TRUE)

# Da .RData (file workspace R, contiene già oggetti nominati)
load("path/to/file.RData")
# dopo load() gli oggetti sono già disponibili con i loro nomi originali

# Da .txt con separatore tab o spazio
df <- read.table("path/to/file.txt", header = TRUE, sep = "\t")
```

## Template — Descriptive statistics for initial data inspection

> Reusable for any exercise that asks to "describe the data" or "report descriptive 
> statistics" as a preliminary inspection (e.g., Es. 5, 10, 11, 12, 13 of the HW).

### Code patterns by data structure

#### Case 1 — Single continuous variable (e.g., Es. 11 sprint times)

```{r desc-single-vec, eval=FALSE}
x <- c(...)            # or x <- df$variable_name

# main descriptive statistics
summary(x)             # min, Q1, median, mean, Q3, max
sd(x)                  # standard deviation
IQR(x)                 # interquartile range
range(x)               # min and max
length(x)              # sample size
```

For a more compact and structured display, use `skimr`:

```{r desc-skimr, eval=FALSE}
require(skimr)
skim_without_charts(data.frame(x = x))
```


```{r, eval=FALSE}
boxplot(df)
plot(density(df))
```

The boxplot of `[variable]` shows a median of approximately `[median]`, 
with the central 50% of the observations concentrated between `[Q1]` and 
`[Q3]` (IQR = `[IQR]`). The box appears `[symmetric / asymmetric]`, with 
the median `[centered / shifted toward Q1 / shifted toward Q3]`.

The whiskers extend from `[lower whisker]` to `[upper whisker]`. `[N]` 
observations fall outside the whiskers and are flagged as potential outliers 
(points above `[upper whisker]` or below `[lower whisker]`). `[Brief 
interpretation: e.g., "these correspond to unusually large/small values 
of [variable]"]`.


The density plot of `[variable]` shows a `[unimodal/bimodal]`, 
`[symmetric/right-skewed/left-skewed]` distribution with mode around 
`[value]`. The shape is `[consistent with / suggests deviation from]` 
a normal distribution, which can be further verified via a Q-Q plot.

#### Case 2 — Two continuous variables (e.g., Es. 12 two servers)

```{r desc-two-vec, eval=FALSE}
A <- c(...)
B <- c(...)

# parallel summaries
summary(A)
summary(B)

# parallel SDs
round(sd(A), 3)
round(sd(B), 3)

# range comparison
diff(range(A))         # spread of A
diff(range(B))         # spread of B
```

#### Case 3 — Multivariate dataset (e.g., Es. 5 school dataset)

```{r desc-multivar, eval=FALSE}
require(skimr)
require(dplyr)

skim_without_charts(df) %>%
  dplyr::select(-complete_rate)

# coefficient of variation (to compare variability across different scales)
cv <- function(x) sd(x) / mean(x)
round(apply(df, 2, cv), 2)
```

#### Case 4 — Continuous variable by group (e.g., bivariate / regression context)

```{r desc-by-group, eval=FALSE}
# summary statistics by group
by(df[, c("var1", "var2")], df$group, summary)

# or with tapply for a single variable
tapply(df$var1, df$group, summary)
```

---

### What to comment — checklist by block

#### Block A — Central tendency

- **Mean vs Median**: are they close or far apart? If similar, distribution is 
  approximately symmetric. If mean > median, positive (right) asymmetry. 
  If mean < median, negative (left) asymmetry.
- The **median** is a more robust measure of the "typical value" than the mean 
  in presence of outliers.

> *Example phrasing*:
> 
> "The mean and median of `[variable]` are very similar to each other ([m] vs 
> [median]), suggesting that the distribution is approximately symmetric around 
> the central value of [value]. Half of the observations fall below [median] 
> [interpretation in context]."

#### Block B — Spread / variability

- **Standard deviation**: typical deviation from the mean. Same units as the
  variable.
- **Interquartile range (IQR)**: spread of the central 50% of the data, 
  robust to outliers.
- **Range** (max - min): total spread. Compare with IQR: if much larger 
  (~ 4-6× IQR), there are extreme values in the tails.
- **Coefficient of variation** ($\sigma/\mu$): (CV)dimensionless measure of 
  variability, useful to compare across variables with different scales. (soglie basso minore 0.1 medio 0.2 alto 0.3)

> *Example phrasing*:
> 
> "The standard deviation of `[variable]` is [SD], indicating that observations 
> deviate on average by approximately [SD] units from the mean. The interquartile 
> range is [IQR], meaning the central 50% of the data are concentrated in a 
> range of [IQR] units. The total range is [range], roughly [range/IQR] times 
> the IQR, [consistent with / suggesting deviation from] a normal-like distribution."

> *When comparing two groups*:
> 
> "The variability of `[variable_A]` is [higher / lower / similar to] that of 
> `[variable_B]` (SD = [SD_A] vs [SD_B]). This suggests that [interpretation in 
> context: e.g., one group is more consistent than the other]."

#### Block C — Quartiles and distribution shape

- **Q1, Q2 (median), Q3**: divide the data into four equal parts.
- Compare $|Q1 - \text{median}|$ vs $|\text{median} - Q3|$:
  - If similar → symmetric distribution
  - If $|\text{median} - Q3| > |Q1 - \text{median}|$ → right skew
  - If $|Q1 - \text{median}| > |\text{median} - Q3|$ → left skew

> *Example phrasing*:
> 
> "The first quartile is [Q1], the third quartile is [Q3]. One-quarter of the 
> observations are below [Q1], another quarter are above [Q3]. The position of 
> the median ([median]) within the IQR suggests [symmetric / right-skewed / 
> left-skewed] distribution."

#### Block D — Extreme values

- **Min and Max**: total range of observed values.
- Compare with the application context: are these values **realistic**? Out-of-scale 
  values may indicate measurement errors or outliers.
- Compute the rule-of-thumb interval $[\bar x - 3s, \bar x + 3s]$ for a normal 
  distribution. Observations outside this interval are unusually extreme.

> *Example phrasing*:
> 
> "The minimum and maximum observed values are [min] and [max] respectively, 
> spanning a range of [range] units. The maximum value [appears / does not appear] 
> realistic given the application context [explain]. Applying the rule of thumb 
> for normal distributions ($\mu \pm 3\sigma$ = [interval]), [N] observations 
> fall outside this interval — [a number consistent with normality / 
> suggesting heavy tails or outliers]."

---

### Generic comment template (combine all blocks)

The descriptive statistics for `[variable_name]` are as follows. The mean is 
[mean] and the median is [median]: their [similarity / difference] suggests 
that the distribution is [approximately symmetric / asymmetric] around the 
central value.

Regarding variability, the standard deviation is [SD], indicating that the 
observations deviate on average by approximately [SD] units from the mean. 
The interquartile range is [IQR], with one-quarter of the observations below 
[Q1] and another quarter above [Q3]. The total range is from [min] to [max], 
approximately [range/IQR] times the IQR, [consistent with / suggesting deviation 
from] a normal-like distribution.

[Optional, in applied contexts]: In substantive terms, these values indicate 
that [interpretation in context — e.g., "the athlete's typical sprint time is 
around 36 seconds, with relatively consistent performance (low IQR), and one 
extreme observation that warrants further investigation"].

[Optional, when comparing groups]: Comparing the two groups, `[group_A]` shows 
[higher / lower / similar] central tendency and [higher / lower / similar] 
variability than `[group_B]`. Specifically, the mean of [variable] is [mean_A] 
in group A versus [mean_B] in group B, and the standard deviation is [SD_A] vs 
[SD_B].

range/iqr: 3-5 compatible 5-7 possbile heavy tails >7-8 deviation from normality/heavy outlier >>10 distribuzione dispersa o heavy tailed

---

### Quick rule of thumb

| Pattern observed | Interpretation |
|------------------|----------------|
| mean ≈ median | Approximately symmetric distribution |
| mean > median | Positive asymmetry (right tail) |
| mean < median | Negative asymmetry (left tail) |
| range ≈ 4-6 × IQR | Consistent with normal distribution |
| range >> 6 × IQR | Heavy tails or outliers present |
| Few observations outside $[\bar x \pm 3s]$ | Consistent with normality |
| Many observations outside $[\bar x \pm 3s]$ | Deviation from normality |
| SD large relative to mean (CV > 1) | High relative variability |
| SD small relative to mean (CV < 0.1) | Highly concentrated data |

---

### Tip — connecting EDA to the next step

Always close the descriptive analysis with a **transition sentence** that 
links to the upcoming analysis. Example endings:

- For regression contexts:
  > "These preliminary observations suggest that [variable] is approximately 
  > normally distributed, supporting the assumption of normality that will be 
  > required in the subsequent regression model. Possible outliers identified 
  > at the descriptive level will be assessed via Cook's distance in the 
  > diagnostic phase."

- For bootstrap contexts:
  > "Given the small sample size and the presence of potential outliers, 
  > the bootstrap procedure is appropriate to evaluate the accuracy of the 
  > estimator without relying on distributional assumptions."

- For bivariate/multivariate contexts:
  > "The univariate descriptives indicate that both variables [appear / do not 
  > appear] consistent with a normal distribution. The bivariate analysis 
  > performed in the next section will further assess the joint distribution 
  > and the association between the two variables."

---

### Practical examples (concrete fill-in)

#### Example 1 — Single variable (Es. 11 sprint times)

After running `summary(times)` and `sd(times)`, you might write:

The descriptive statistics for the 200m sprint times show a median of [36.31] 
seconds and a mean of [36.94] seconds. The slight difference between mean and 
median (with mean higher) suggests a positive asymmetry in the distribution.

The interquartile range is [1.63] seconds (Q1 = [35.50], Q3 = [37.13]), 
indicating that the central 50% of times are concentrated in a narrow window, 
consistent with the athlete's relative performance consistency.

However, the maximum observed value ([84.07] seconds) is noticeably distant 
from Q3 (approximately [38] times the IQR), suggesting the presence of an 
outlier — possibly due to a measurement error or a non-typical event during 
that specific trial. The minimum value ([22.48] seconds) appears realistic 
as a personal best.

#### Example 2 — Two groups (Es. 12 servers)

After running `summary(A)` and `summary(B)`, you might write:

The descriptive statistics show that server B is consistently faster than 
server A for the eight processed files. The mean processing time is [158.9] 
ms for server A versus [146.9] ms for server B — a difference of approximately 
[12] ms on average. Similarly, the median is [159.5] ms for A and [144.5] ms 
for B, confirming the systematic advantage of server B.

In terms of variability, the standard deviation is [17.2] ms for server A and 
[15.6] ms for server B: server A shows slightly more variable performance, 
although the difference is modest. The range of values is comparable between 
the two servers ([55] ms for A vs [48] ms for B).

Overall, the descriptive analysis suggests that server B not only performs 
faster on average, but also with marginally higher consistency. Whether this 
difference is statistically significant given the small sample size (n = 8) 
will be assessed via the bootstrap procedure in the next section.

### Conversione di tipo (frequente)

```{r type-conversion, eval=FALSE}
df$variabile_cat <- as.factor(df$variabile_cat)    # numerica → categorica
df$variabile_num <- as.numeric(df$variabile_num)   # carattere → numerica
levels(df$variabile_cat)                           # mostra i livelli del factor
```
---

# Distribuzioni e Variabili Casuali

## Distribuzioni e variabili casuali

### Le 4 funzioni R per ogni distribuzione

Per ogni distribuzione, R fornisce 4 funzioni con prefisso fisso:

| Prefisso | Cosa fa | Input → Output |
|----------|---------|----------------|
| `d` | densità (pdf) o massa (pmf) in un punto | $x$ → $f(x)$ |
| `p` | probabilità cumulata $P(X \le x)$ | $x$ → $F(x)$ |
| `q` | quantile (inverso di p) | $p \in [0,1]$ → $x$ tale che $F(x) = p$ |
| `r` | genera $n$ valori casuali | $n$ → vettore di realizzazioni |

Distribuzioni rilevanti per il corso:

| Distribuzione | Suffisso | Parametri |
|---------------|----------|-----------|
| Normale | `norm` | `mean`, `sd` |
| Student-t | `t` | `df` |
| Chi-quadro | `chisq` | `df` |
| F di Fisher | `f` | `df1`, `df2` |
| Binomiale | `binom` | `size`, `prob` |

Esempi:

```{r dist-functions, eval=FALSE}
# NORMALE N(140, 6^2)
dnorm(135, mean = 140, sd = 6)        # densità f(135)
pnorm(130, mean = 140, sd = 6)        # P(X <= 130) ≈ 0.0478
qnorm(0.975, mean = 140, sd = 6)      # 97.5-esimo percentile ≈ 151.76
rnorm(1000, mean = 140, sd = 6)       # 1000 realizzazioni

# STUDENT-T con 10 df
qt(0.975, df = 10)                     # 97.5-esimo percentile (uso tipico: CI)
pt(2.5, df = 10)                       # P(T <= 2.5)

# Quantili tipici per intervalli di confidenza:
qnorm(0.975)                           # ≈ 1.96, per CI al 95% con sigma noto
qt(0.975, df = n - 1)                  # per CI al 95% con sigma stimato
```

**Comment template (typical use in tests/CIs):**
> The theoretical 97.5th percentile of the [normal / t with `df` degrees of freedom] distribution is `[value]`. I use it to construct a 95% confidence interval for `[parameter]`.

### Generare campioni e trasformazioni lineari

```{r generate-samples, eval=FALSE}
# Riproducibilità: SEMPRE prima di rnorm/rt/sample/...
set.seed(7369)

# 1000 valori da N(0, 1)
z <- rnorm(1000, mean = 0, sd = 1)

# trasformazione lineare: Y = mu + sigma * Z  ⇒  Y ~ N(mu, sigma^2)
y <- 140 + 6 * z          # equivalente a rnorm(1000, 140, 6)

# campione da una t-Student con 3 df (code pesanti)
t_vals <- rt(1000, df = 3)
```

**Concetto chiave**: una normale generale $\mathcal{N}(\mu, \sigma^2)$ si ottiene **standardizzando alla rovescia** una standard $\mathcal{N}(0,1)$:
$$Y = \mu + \sigma Z$$

E viceversa, per **standardizzare**:
$$Z = \frac{Y - \mu}{\sigma} \sim \mathcal{N}(0, 1)$$

```{r standardize, eval=FALSE}
# standardizzare un vettore osservato
z_std <- (y - mean(y)) / sd(y)         # standardizzazione campionaria
# oppure:
z_std <- scale(y)                       # funzione R built-in
```

**Quando standardizzare:**
- Per confrontare variabili su scale diverse
- Prima di alcune procedure (es. clustering, PCA)
- Per calcolare quantili teorici

**Note sul `set.seed()`:**
- Va messo **subito prima** di ogni funzione che genera casualità
- Stessa seed → stessi numeri casuali → risultati riproducibili
- Cambiare seed cambia il campione, ma non le proprietà generali del campione (per `n` grande)

### Visualizzazione: istogramma e ECDF

#### Istogramma

```{r hist, eval=FALSE}
hist(y,
     breaks = 50,                     # numero di classi (default: regola di Sturges)
     prob   = TRUE,                   # densità invece di frequenze assolute
     main   = "Realizzazioni da Y ~ N(mu, sigma^2)",
     xlab   = "y",
     ylab   = "Densità",
     col    = "lightblue")
```

**Cosa cercare:**

- **Forma**: campanulare (normale) / asimmetrica / multimodale?
- **Centro**: dove si trova il picco?
- **Dispersione**: distribuzione larga o concentrata?

#### Confronto di due distribuzioni

```{r hist-compare, eval=FALSE}
par(mfrow = c(1, 2))                  # 1 riga, 2 colonne di plot

hist(y,  breaks = 50, prob = TRUE,
     ylim = c(0, 0.08), xlim = c(100, 190),
     col  = "yellow1",
     main = "Y ~ N(140, 36)", ylab = "Densità")

hist(y1, breaks = 50, prob = TRUE,
     ylim = c(0, 0.08), xlim = c(100, 190),
     col  = "springgreen2",
     main = "Y1 ~ N(140, 100)", ylab = "Densità")

par(mfrow = c(1, 1))                  # ripristina il default
```

> **Trucco**: per confrontare distribuzioni serve **fissare gli stessi `xlim` e `ylim`** in entrambi gli istogrammi, altrimenti l'occhio inganna.

#### ECDF (empirical cumulative distribution function)

Stima non parametrica della cdf vera, basata sulla frazione di dati $\le x$.

```{r ecdf, eval=FALSE}
plot(ecdf(y),
     main = "ECDF a confronto",
     xlab = "x",
     ylab = expression(F[n](x)))
plot(ecdf(y1), add = TRUE, col = "red")
legend("topleft",
       legend = c("N(140, 36)", "N(140, 100)"),
       lty = c(1, 1), lwd = c(2, 2),
       col = c("black", "red"),
       cex = 0.8)
```

**Come si legge:**

- Funzione a gradini, ogni gradino = un'osservazione
- Asse y: frazione di dati $\le x$. Va da 0 a 1
- **Curva più ripida al centro** = distribuzione più concentrata (varianza minore)
- **Curva più dolce** = distribuzione più dispersa

**Comment template:**

The empirical cumulative distribution function (ECDF) of the variable 
**[NAME]** estimates non-parametrically the underlying distribution by 
reporting, for each value $x$, the proportion of observations not 
exceeding it. By construction it is a non-decreasing step function valued 
in $[0, 1]$, with jumps of size $1/n$ at each observed data point, and 
by the Glivenko–Cantelli theorem it converges uniformly to the 
theoretical CDF as $n$ grows.

From the plot we observe that the ECDF is [steep around the central 
values / gently sloped over a wide range], indicating a distribution 
[concentrated / spread out] around [MEDIAN]. The curve crosses the level 
$0.5$ at approximately $x = $ [MEDIAN], while the levels $0.25$ and 
$0.75$ are reached at [Q1] and [Q3], giving an interquartile range of 
[IQR]. The lower and upper tails approach $0$ and $1$ 
[rapidly / gradually], suggesting [few / several] extreme values, and 
the overall shape appears [symmetric / right-skewed / left-skewed] 
around the median.


IQR≪σ⋅1,35 (cioè IQR molto più piccolo del "IQR atteso" di una normale con quella σ) → ECDF molto steep al centro, code lunghe
IQR≈1,35⋅σ → distribuzione circa normale, ECDF con la classica "S" simmetrica
IQR≫1,35⋅σ → dati spalmati, ECDF gently sloped
---

### QQ-plot

Confronto visivo tra distribuzione empirica e teorica (di solito normale). È lo strumento principe per la diagnostica della normalità — userai questo per i residui della regressione.

#### Costruzione

```{r qqplot, eval=FALSE}
qqnorm(y, col = "blue", main = "QQ-plot di y vs Normale")
qqline(y)                              # retta di riferimento
```

#### Confronto: normale vs t con code pesanti

```{r qqplot-compare, eval=FALSE}
set.seed(123)
Y1 <- rnorm(1000)                      # normale standard
set.seed(234)
Y2 <- rt(1000, df = 3)                 # t-Student con 3 df

par(mfrow = c(1, 2))
qqnorm(Y1, col = "blue", main = "Y1 ~ N(0,1)"); qqline(Y1)
qqnorm(Y2, col = "blue", main = "Y2 ~ t(3)");   qqline(Y2)
par(mfrow = c(1, 1))
```

#### Tabella di interpretazione (cruciale)

| Pattern del QQ-plot | Diagnosi |
|---------------------|----------|
| Punti allineati sulla retta | Distribuzione **normale** ✓ |
| Curva a "S" con estremi che salgono/scendono lontano dalla retta | **Code pesanti** (più valori estremi del previsto) |
| Concavità verso l'alto su tutto il range | **Asimmetria positiva** (coda destra lunga) |
| Concavità verso il basso | **Asimmetria negativa** (coda sinistra lunga) |
| Pochi punti isolati lontano dalla retta agli estremi | **Outlier** |

**Comment template (exam):**

The quantile-quantile (QQ) plot compares the empirical quantiles of the 
variable **[NAME]** with the theoretical quantiles of a reference 
distribution (typically the standard normal): if the data follow the 
assumed model, the points should lie approximately along the reference 
line, while systematic departures from linearity reveal discrepancies 
between the empirical and theoretical distributions.

From the plot we observe that the points lie [closely along / 
systematically off] the reference line, indicating that the distribution 
of **[NAME]** is [well approximated by / not fully compatible with] the 
[Normal / theoretical] distribution. In particular, the central portion 
of the plot appears [aligned with / slightly off] the line, suggesting 
that the bulk of the data is [consistent / not consistent] with the 
reference model, while the [lower / upper / both] tail(s) show 
[no noticeable deviation / a clear deviation], with points falling 
[above / below] the line. This behavior is typical of a 
[symmetric / right-skewed / left-skewed / heavy-tailed / light-tailed] 
distribution. Overall, the QQ-plot suggests that the [Normal / theoretical] 
assumption is [reasonable / questionable] for **[NAME]**, and this visual 
assessment can be complemented by formal goodness-of-fit tests such as 
Shapiro–Wilk test.

#### Test formali di normalità (alternativa al QQ-plot)

```{r normality-tests, eval=FALSE}
shapiro.test(y)                        # test di Shapiro-Wilk (n piccolo, n < 5000)
# H0: i dati provengono da una normale
# p-value piccolo → rifiuto la normalità
```

> **Attenzione**: con `n` molto grande, anche piccole deviazioni dalla normalità portano a rifiutare H0. Il QQ-plot rimane più informativo del test in molti casi pratici.

## Normale bivariata e multivariata
Normale bivariata e multivariata
 
### Generazione di realizzazioni
 
```{r mvnorm-generate, eval=FALSE}
library(mvtnorm)
 
set.seed(625)
 
# vettore delle medie
mu <- c(0, 0)
 
# matrice di varianza-covarianza
# diagonale = varianze (sigma^2), fuori diagonale = covarianze
sigma <- matrix(c(9, 2,
                  2, 9), ncol = 2)
 
# generazione
X <- rmvnorm(n = 3000, mean = mu, sigma = sigma)
# X è una matrice n × 2: una riga per realizzazione, una colonna per componente
#se bivariate con stessa varianza - stesso iqr
```
 
> **Attenzione alla parametrizzazione**: nella matrice `sigma` la diagonale contiene le **varianze**, non le deviazioni standard. Se ti viene dato $\sigma = 3$, devi inserire $\sigma^2 = 9$ in `sigma`.
 
### Costruzione di Sigma a partire da SD e correlazione
 
Quando ti vengono dati SD e correlazioni invece di varianze e covarianze:
 
```{r sigma-from-corr, eval=FALSE}
sd_vec <- c(3, 3)              # deviazioni standard
R <- matrix(c(1.0, 0.7,        # matrice di correlazione
              0.7, 1.0), ncol = 2)
 
D <- diag(sd_vec)
sigma <- D %*% R %*% D         # Sigma = D R D
```

### Descriptive statistics
```{r desc, eval=FALSE}
summary(X)
sd(X[, 1])

sd(X[, 2])

cov(X[, 1], X[, 2])
```

### Descriptive Statistics Comment Template — summary() and sd()

The object returned in the output by the [FUNCTION_NAME] function is a 
matrix with two columns and [N_OBS] rows: the two columns represent the two 
components ([VAR_X] and [VAR_Y]) of the bivariate variable, while each row 
corresponds to a different observation. We briefly analyze the generated 
data by calculating its main descriptive statistics. Firstly, we recall 
that the components of a bivariate Gaussian distributed random variable 
have univariate Gaussian distributions; in this case, given the assigned 
parameters, we expect [VAR_X] ~ N([MU_X], [VAR_X_VAL]) and 
[VAR_Y] ~ N([MU_Y], [VAR_Y_VAL]). From the descriptive statistics we 
observe that, for both univariate components, the mean and median assume 
very similar values to each other and are approximately equal to [MU_X] 
and [MU_Y] respectively, which correspond to the theoretical values 
assigned to the means. The study of the quartiles also highlights the 
[symmetry/asymmetry] of the data around the central values of the mean 
and median, since the distance between Q1 and the median is 
[comparable to / noticeably different from] the distance between the 
median and Q3 for both components. In summary, the values generated for 
both components appear to be [consistent / not fully consistent] with 
those of a random variable having the required Gaussian distribution.

The standard deviations of the two components, equal to sd([VAR_X]) = 
[SD_X] and sd([VAR_Y]) = [SD_Y], are both very close to the assigned 
theoretical values of [SD_X_TH] and [SD_Y_TH]; also in this case, the 
values of the indices calculated on the simulated data are consistent 
with the theoretical ones. Finally, the value of the covariance 
(equal to [COV_XY]) is [slightly greater than / slightly smaller than / 
very close to] the theoretical one ([COV_XY_TH]), and the corresponding 
correlation coefficient ρ = [RHO] confirms that the two components are 
[strongly / moderately / very weakly] linearly and 
[positively/negatively] associated with each other.

### Basic scatterplot
```{r, eval=FALSE}
plot(X[, 1], X[, 2], col = "blue", main = "Scatter plot",
xlab = "First component (X)", ylab = "Second component (Y)")
```

### Template — Bivariate scatterplot (2D, Gaussian context)

The scatter plot shows the realizations of the bivariate `[Gaussian / 
empirical]` distribution in the Cartesian plane, with `[X_1]` on the 
horizontal axis and `[X_2]` on the vertical axis.

Most of the points are clustered around the mean `[(μ_X1, μ_X2)]`, with a 
`[limited / substantial]` number of points that deviate from the central 
part of the distribution. The range of variation is `[approximately the 
same / different]` for the two components, indicating that the corresponding 
variances are `[similar / different]`. 

The points are arranged in a region of `[circular / elliptical]` shape, 
`[without / with]` a `[well-defined / clear]` orientation `[along the "/" 
direction / along the "\\" direction]`. This behaviour indicates that the 
covariance is `[equal to zero / positive / negative]`.

[Optional, if outliers or non-typical observations are present:]

The plot also reveals the presence of `[describe: e.g., "a second, less 
dense cloud of points characterized by high values of X_1" or "an isolated 
point (outlier) showing the highest values for both X_1 and X_2"]`, which 
will require attention in the subsequent analysis.
Guarda l’orientamento della “nuvola”:

inclinata verso l’alto → correlazione positiva
inclinata verso il basso → correlazione negativa
quasi circolare → correlazione circa zero

Più la nuvola è “stretta” e allungata → correlazione forte.

Più è “larga” e quasi circolare → correlazione debole.

### Densità teorica e plot 3D
 
```{r density-3d, eval=FALSE}
library(scatterplot3d)
 
z1 <- seq(min(X[, 1]), max(X[, 1]), length.out = 50)
z2 <- seq(min(X[, 2]), max(X[, 2]), length.out = 50)
grid <- expand.grid(z1, z2)
dens <- matrix(dmvnorm(grid, mean = mu, sigma = sigma),
               ncol = length(z1), byrow = TRUE)
 
plot3d <- scatterplot3d(z1, z2,
                        seq(min(dens), max(dens), length.out = length(z1)),
                        type = "n", angle = 70, grid = FALSE,
                        main = "Densità teorica bivariata",
                        xlab = expression(X[1]), ylab = expression(X[2]),
                        zlab = "f(X1, X2)")
 
for (i in length(z1):1) {
  plot3d$points3d(rep(z1[i], length(z2)), z2, dens[i, ],
                  type = "l", col = "darkorange")
}
for (i in length(z2):1) {
  plot3d$points3d(z1, rep(z2[i], length(z1)), dens[, i],
                  type = "l", col = "darkorange")
}
```

The scatter3d shows the bivariate normal density of **[VAR_X]** and 
**[VAR_Y]** with μ = ([MU_X], [MU_Y]) and correlation ρ = [RHO]. The surface 
peaks at the mean and decays smoothly in every direction. The elliptical 
contours are [tilted positively / tilted negatively / axis-aligned], 
indicating a [positive / negative / null] linear dependence between the two 
variables, while the [VAR_X / VAR_Y] axis shows [greater / smaller] spread 
due to σ_x [>/</≈] σ_y.

### Contour plot (ellissi di isodensità)
 
```{r contour, eval=FALSE}
z1 <- seq(-10, 10, length.out = 300)
z2 <- seq(-10, 10, length.out = 300)
grid <- expand.grid(z1, z2)
dens <- matrix(dmvnorm(grid, mean = mu, sigma = sigma), ncol = length(z1))
 
contour(z1, z2, dens,
        col = "blue",
        main = "Contour plot",
        xlab = expression(X[1]), ylab = expression(X[2]),
        asp = 1)                     # asp=1 mantiene le proporzioni!
```
 
### Tabella di interpretazione delle ellissi
 
| Forma osservata | Diagnosi |
|-----------------|----------|
| Cerchio perfetto | $\rho = 0$ **e** $\sigma_{X_1} = \sigma_{X_2}$ |
| Ellisse con assi orizzontale/verticale | $\rho = 0$, $\sigma_{X_1} \neq \sigma_{X_2}$ |
| Ellisse inclinata "/" | $\rho > 0$ |
| Ellisse inclinata "\\" | $\rho < 0$ |
| Ellisse molto schiacciata | $|\rho|$ vicino a 1 |

Inclinazione (segno di ρ)
Guarda la direzione di massima estensione (cresta nel 3D, asse maggiore dell'ellisse nel contour):
↗ da basso-sx ad alto-dx → ρ > 0
↘ da alto-sx a basso-dx → ρ < 0
∥ asse X o Y (orizzontale/verticale) → ρ ≈ 0
Forma rotonda/cerchio/collina simmetrica → ρ ≈ 0 e σ_x ≈ σ_y

Step 3 — Rapporto σ_x / σ_y
Guarda la proiezione sugli assi (l'ombra dell'ellisse o della base della campana):

Si estende di più lungo X → σ_x > σ_y
Si estende di più lungo Y → σ_y > σ_x
Proiezioni simili su entrambi → σ_x ≈ σ_y

Step 4 — Solo per il 3D: altezza del picco

Campana alta e stretta → varianze piccole e/o |ρ| grande → probabilità concentrata
Campana bassa e larga → varianze grandi e/o ρ ≈ 0 → probabilità diffusa

The contour plot displays the level sets of the bivariate normal density of 
**[VAR_X]** and **[VAR_Y]**, with mean vector μ = ([MU_X], [MU_Y]) and 
correlation ρ = [RHO]. Each curve connects points of equal density, providing 
a 2D projection of the underlying 3D surface.

The contours are **elliptical** and concentric around the mean ([MU_X], 
[MU_Y]), which corresponds to the mode of the distribution. Their geometry 
reflects the covariance structure:

- **Orientation:** the ellipses are [tilted along the positive diagonal / 
  tilted along the negative diagonal / aligned with the coordinate axes], 
  consistent with a [positive / negative / negligible] correlation 
  ρ = [RHO].
- **Eccentricity:** the ellipses are [highly elongated / moderately 
  elongated / nearly circular], indicating that the linear dependence between 
  [VAR_X] and [VAR_Y] is [strong / moderate / weak]. As |ρ| → 1 the contours 
  collapse onto a line; as ρ → 0 they become axis-aligned ellipses (or 
  circles when σ_x = σ_y).
- **Axis spread:** the ellipses extend [further along VAR_X / further along 
  VAR_Y / similarly along both axes], reflecting σ_x [>/</≈] σ_y.

The innermost contour identifies the region of highest probability mass, 
while outer contours enclose increasingly larger probability regions (e.g. 
the ellipse at level corresponding to χ²₂(0.95) contains approximately 95% 
of the distribution).
 
### Decomposizione spettrale di $\Sigma$
 
```{r spectral, eval=FALSE}
eig <- eigen(sigma)
eig$values        # autovalori = varianze lungo gli assi principali
eig$vectors       # autovettori = direzioni degli assi principali
 
# verifica della decomposizione: Sigma = Q Lambda Q^T
eig$vectors %*% diag(eig$values) %*% t(eig$vectors)
```
 
**Comment template (typical use):**

> The contour plot shows [circular / tilted / flattened] ellipses, consistent with [the parameter values of the distribution]. The shape [is / is not] consistent with what is expected under bivariate normality: in particular, [observed shape] suggests that $\rho$ is [close to zero / positive / negative] and that the variances [are / are not] of the same order of magnitude.

### Proprietà chiave (mappe mentali per il commento)

- **Marginali**: di una bivariata normale sono univariate normali. Verificale con `qqnorm()` su ciascuna colonna.
- **Indipendenza** equivale a ρ = 0 **solo per la normale** (non vale in generale).
- **Media condizionata**: E(Y|X=x) è lineare in x. È il fondamento della regressione lineare.
- **Varianza condizionata**: Var(Y|X=x) = σ²_Y · (1 - ρ²). Conoscere X riduce l'incertezza su Y di un fattore (1 - ρ²).
---
 
## Misure di associazione
 
### Covarianza e correlazione
 
```{r cov-cor, eval=FALSE}
# tra due variabili
cov(df$x, df$y)
cor(df$x, df$y)
 
# matrice di var-cov e di correlazione (per più variabili)
cov(df)
cor(df)
round(cor(df), 2)              # arrotondata per leggibilità
```
 
**Differenza pratica:**
 
- `cov` dipende dalle unità di misura → leggibile solo se confronti con SD
- `cor` è in $[-1, 1]$ → leggibile direttamente
**Soglie indicative per $|\rho|$:**
 
| $|\rho|$ | Forza |
|----------|-------|
| < 0.2 | Debole |
| 0.2 – 0.5 | Moderata |
| 0.5 – 0.7 | Significativa |
| 0.7 – 0.9 | Forte |
| > 0.9 | Molto forte (attenzione a multicollinearità) |
 
### Correlazione parziale
 
Risolve il problema dei **confondenti**: misura l'associazione tra X e Y al netto di Z.
 
```{r partial-cor, eval=FALSE}
library(ppcor)
 
# correlazione parziale tra due variabili al netto di tutte le altre
pcor(df)$estimate              # matrice delle correlazioni parziali
pcor(df)$p.value               # p-value associati
 
# tra coppia specifica al netto di una variabile specifica
pcor.test(x = df$x, y = df$y, z = df$z)
```
 
**Interpretazione (i 3 casi tipici):**
 
| $r_{xy}$ | $r_{xy \cdot z}$ | Diagnosi |
|----------|------------------|----------|
| Alta | Vicina a 0 | **Correlazione spuria**: l'associazione era spiegata da Z |
| Alta | Ancora alta | **Associazione genuina**: persiste anche al netto di Z |
| Vicina a 0 | Alta | **Paradosso di Simpson**: Z mascherava un'associazione reale |
 
**Comment template:**

> The marginal correlation between `[X]` and `[Y]` is `[value]`, while the partial correlation controlling for `[Z]` is `[value]`. The [reduction / persistence / reversal] suggests that [the association is explained by Z / the association is genuine / Z was masking the true relationship].
 
### Scatterplot e scatterplot matrix
 
```{r scatter-base, eval=FALSE}
# scatterplot bivariato
plot(df$x, df$y,
     pch = 16, cex = 0.8, col = "blue",
     main = "Y vs X",
     xlab = "X", ylab = "Y")
abline(lm(y ~ x, data = df), col = "red")    # retta di regressione opzionale
 
# scatterplot matrix (base)
pairs(df, panel = panel.smooth, main = "Scatterplot matrix")
 
# versione informativa con GGally
library(GGally)
ggpairs(df)
# diagonale: densità marginali
# sotto diagonale: scatterplot
# sopra diagonale: correlazioni con livelli di significatività
```


### Visualizzazione matrici di correlazione: corrplot
 
```{r corrplot, eval=FALSE}
library(corrplot)
 
# correlazioni grezze
corrplot(cor(df),
         type = "upper",                   # solo triangolo superiore
         method = "circle",                # circle / square / number / color
         tl.col = "black",                 # colore etichette
         tl.cex = 0.8)                     # dimensione etichette
 
# correlazioni parziali
corrplot(pcor(df)$estimate,
         type = "upper",
         method = "circle",
         tl.col = "black")
 
# affiancate per confronto visivo
par(mfrow = c(1, 2))
corrplot(cor(df), type = "upper", method = "circle", title = "Raw")
corrplot(pcor(df)$estimate, type = "upper", method = "circle", title = "Partial")
par(mfrow = c(1, 1))
```
 
> **Trucco**: `corrplot` affiancati di correlazioni *raw* vs *partial* è il modo più rapido per individuare confondenti e relazioni spurie a colpo d'occhio.
 
### Trappole comuni nelle misure di associazione
 
- **$\rho = 0$ non implica indipendenza** (tranne sotto normalità bivariata). Esempio classico: $Y = X^2$ con $X$ simmetrica → $\rho = 0$ ma forte dipendenza
- **Correlazione lineare** è cieca alle relazioni non lineari → guarda sempre lo scatterplot prima di fidarti di $\rho$
- **Correlazione $\neq$ causazione** (esempio gelati-annegamenti) → la correlazione parziale aiuta a capire se l'associazione è genuina, ma non basta da sola per stabilire causalità (servirebbe un esperimento randomizzato)
- **Multicollinearità**: se $|\rho| > 0.9$ tra due covariate, la regressione successiva avrà problemi (lo affronteremo con VIF)


# Inferenza

## Estimators properties

An estimator $\hat\theta$ is a statistic, that is, a function of the sample 
$(X_1, \dots, X_n)$, used to approximate an unknown parameter $\theta$ of 
the population distribution. Since different estimators can be proposed for 
the same parameter, it is essential to identify the properties that 
distinguish a "good" estimator from a poor one. These properties are 
usually divided into **finite-sample properties**, which hold for any 
sample size $n$, and **asymptotic properties**, which hold in the limit as 
$n \to \infty$.

---

**1. Unbiasedness.** An estimator $\hat\theta$ is said to be *unbiased* if 
its expected value coincides with the true parameter:

$$
\mathbb{E}[\hat\theta] = \theta.
$$

The *bias* of the estimator is defined as 
$\mathrm{Bias}(\hat\theta) = \mathbb{E}[\hat\theta] - \theta$, and 
unbiasedness corresponds to $\mathrm{Bias}(\hat\theta) = 0$. This property 
is important because it guarantees that, on average over repeated samples, 
the estimator neither systematically overestimates nor underestimates the 
parameter. A biased estimator can still be useful, but only if its bias is 
small or vanishes asymptotically.

**2. Efficiency (minimum variance).** Among all unbiased estimators of 
$\theta$, the most desirable is the one with the smallest variance, since 
a smaller variance means that the estimates are more concentrated around 
the true value $\theta$ and therefore more reliable. An unbiased estimator 
whose variance attains the **Cramér–Rao lower bound** 

$$
\mathrm{Var}(\hat\theta) \;\geq\; \frac{1}{I(\theta)},
$$ 

where $I(\theta)$ denotes the Fisher information, is called *efficient*. 
Efficiency is crucial because it ensures that no other unbiased estimator 
can exploit the information contained in the sample more effectively.

**3. Mean Squared Error (MSE).** When comparing estimators that are not 
both unbiased, the natural criterion is the *mean squared error*, which 
combines bias and variance:

$$
\mathrm{MSE}(\hat\theta) 
= \mathbb{E}\!\left[(\hat\theta - \theta)^2\right] 
= \mathrm{Var}(\hat\theta) + \mathrm{Bias}(\hat\theta)^2.
$$

This decomposition highlights the well-known *bias–variance trade-off*: a 
slightly biased estimator with a much smaller variance can outperform an 
unbiased one with high variability. The MSE is therefore the most 
comprehensive single measure of the quality of an estimator in finite 
samples.

**4. Consistency.** An estimator $\hat\theta_n$ is said to be *consistent* 
if it converges in probability to the true value of the parameter as the 
sample size grows:

$$
\hat\theta_n \;\xrightarrow{P}\; \theta 
\qquad \text{as } n \to \infty.
$$

Consistency is a minimal requirement for any reasonable estimator: it 
guarantees that, by collecting enough data, we can approximate the true 
parameter with arbitrarily high accuracy. An inconsistent estimator is 
generally regarded as unacceptable, since increasing the sample size would 
not improve it.

**5. Asymptotic unbiasedness.** A weaker condition than unbiasedness is 
*asymptotic unbiasedness*, which requires that 
$\mathbb{E}[\hat\theta_n] \to \theta$ as $n \to \infty$. Many estimators 
(including the maximum likelihood estimator in general) are biased for 
finite $n$ but become unbiased in the limit. This property is important 
because it ensures that the bias becomes negligible for sufficiently large 
samples.

**6. Asymptotic normality.** An estimator is *asymptotically normal* if 
its standardized version converges in distribution to a Gaussian:

$$
\sqrt{n}\,(\hat\theta_n - \theta) 
\;\xrightarrow{d}\; \mathcal{N}\!\left(0,\; V(\theta)\right),
$$

so that for large $n$ we can approximate 
$\hat\theta_n \approx \mathcal{N}(\theta, V(\theta)/n)$. This property is 
fundamental in practice because it provides the theoretical basis for 
constructing **confidence intervals** and performing **hypothesis tests** 
(typically of Wald type) without knowing the exact finite-sample 
distribution of the estimator.

**7. Asymptotic efficiency.** An estimator is *asymptotically efficient* 
if, among all consistent and asymptotically normal estimators, it attains 
the smallest possible asymptotic variance, namely the Cramér–Rao lower 
bound $1/I(\theta)$. This is the asymptotic counterpart of efficiency and 
is one of the main reasons why maximum likelihood estimators are so 
widely used: under standard regularity conditions they are consistent, 
asymptotically unbiased, asymptotically normal and asymptotically 
efficient.

**8. Sufficiency.** An estimator (or, more generally, a statistic) is 
said to be *sufficient* for $\theta$ if it contains all the information 
about $\theta$ that is present in the sample, in the sense that the 
conditional distribution of the data given the sufficient statistic does 
not depend on $\theta$. Sufficiency is important because it implies that 
no information is lost by summarizing the sample through the statistic, 
and by the Rao–Blackwell theorem any unbiased estimator can be improved 
(in terms of variance) by conditioning on a sufficient statistic.


## Maximum Likelihood Estimation

### Concetti chiave

- **Likelihood** $L(\theta) = \prod_i f(x_i; \theta)$: funzione di $\theta$ dati i dati osservati. Misura quanto un valore di $\theta$ "spiega bene" i dati.
- **Log-likelihood** $\ell(\theta) = \sum_i \log f(x_i; \theta)$: si usa sempre questa (più stabile numericamente, derivate più semplici, stesso punto di massimo).
- **MLE**: $\hat\theta_{ML} = \arg\max \ell(\theta)$.

**Proprietà degli stimatori MLE** (utili da citare nei commenti):

- **Consistente** (converge a $\theta$ per $n \to \infty$)
- **Asintoticamente unbiased**
- **Asintoticamente normale**: $\hat\theta \sim \mathcal{N}(\theta, 1/I(\theta))$
- **Asintoticamente efficiente** (varianza minima tra stimatori consistenti)

La **Fisher information** $I(\theta)$ misura la curvatura della log-likelihood al massimo. Più curvatura significa più informazione e quindi standard error più piccoli.

### Esempi standard (formule chiuse)

Quando esistono in forma chiusa, basta calcolarle direttamente:

| Modello | Parametro | MLE |
|---------|-----------|-----|
| Bernoulli($p$) | $p$ | $\hat p = \bar x$ |
| Binomiale($n, p$) | $p$ | $\hat p = x/n$ |
| $\mathcal{N}(\mu, \sigma^2)$, $\sigma$ noto | $\mu$ | $\hat\mu = \bar x$ |
| $\mathcal{N}(\mu, \sigma^2)$ entrambi ignoti | $\mu, \sigma^2$ | $\hat\mu = \bar x$, $\hat\sigma^2 = \frac{1}{n}\sum(x_i - \bar x)^2$ |
| Poisson($\lambda$) | $\lambda$ | $\hat\lambda = \bar x$ |

### Pattern: log-likelihood + ricerca su griglia

Adattabile a qualunque modello sostituendo `dXXX`:

```{r mle-grid, eval=FALSE}
# log-likelihood Bernoulli (esempio del prof)
logLik <- function(p, data) {
  sum(data * log(p) + (1 - data) * log(1 - p))
}

# oppure più generico, usando dnorm/dbinom/dpois con log = TRUE
logLik <- function(theta, data) {
  sum(dnorm(data, mean = theta, sd = sigma_known, log = TRUE))
}

# valuto su griglia
theta_vals <- seq(theta_min, theta_max, length.out = 200)
loglik_vals <- sapply(theta_vals, logLik, data = my_data)

# MLE numerica
theta_mle <- theta_vals[which.max(loglik_vals)]
```

### Pattern: ottimizzazione numerica con `optim()`

Gold standard per problemi multiparametrici o senza formula chiusa:

```{r mle-optim, eval=FALSE}
# negazione perché optim minimizza
neg_logLik <- function(theta, data) {
  -sum(dXXX(data, theta, log = TRUE))    # dXXX = dnorm, dbinom, ...
}

# 1-D: usare metodo Brent con bound espliciti
result <- optim(par    = valore_iniziale,
                fn     = neg_logLik,
                data   = my_data,
                method = "Brent",
                lower  = lower_bound,
                upper  = upper_bound)

result$par              # MLE
result$value            # valore di -logLik al massimo (negativo della max log-lik)

# Multidim: usare BFGS o Nelder-Mead (default)
result <- optim(par    = c(theta1_init, theta2_init),
                fn     = neg_logLik,
                data   = my_data,
                method = "BFGS")
```

### Visualizzazione log-likelihood

```{r mle-plot, eval=FALSE}
plot(theta_vals, loglik_vals,
     type = "l", lwd = 2, col = "blue",
     main = "Log-likelihood",
     xlab = expression(theta),
     ylab = expression(log~L(theta)))

abline(v = theta_mle, col = "red", lty = 2, lwd = 2)
```

### Comment template (exam)

> To estimate the parameter [name], I apply the maximum likelihood method. The log-likelihood, evaluated on a grid of values in [range], has a unique maximum at $\hat\theta = $ [value]. The numerical result [coincides with the analytical estimate $\hat\theta = ...$ / matches the estimate from `optim()`].
>
> The log-likelihood appears [flat / concentrated] around the maximum, indicating a [low-precision / precise] estimate. By the asymptotic normality property, $\hat\theta \sim \mathcal{N}(\theta, 1/I(\theta))$: the curvature of the log-likelihood at the maximum determines the standard error of the estimate.

### Trappole comuni

- `optim()` **minimizza** per default → passare `-logLik` (la funzione negata)
- Per problemi 1-D con vincoli (es. $p \in (0,1)$, $\sigma > 0$), usare `method = "Brent"` con `lower` e `upper` espliciti
- Usare `dnorm(..., log = TRUE)` invece di `log(dnorm(...))`: più stabile numericamente
- Il `set.seed()` non serve nel MLE in sé: serve solo se generi i dati simulati prima

---

## Bootstrap e intervalli di confidenza

### Concetti chiave

- **Idea**: simulare la variabilità di $\hat\theta$ ricampionando dal campione originale **con ripetizione**.
- **Plug-in principle**: $\hat F_n$ (ECDF) sostituisce $F$ (popolazione ignota).
- **Procedura in 4 passi**: resample → compute → repeat $B$ volte → usa la distribuzione empirica.

**Quanti $B$?**

- $B = 200$ basta per stimare lo standard error
- $B = 1000$–$2000$ raccomandato per costruire intervalli di confidenza

**Quando usarlo**: stimatori complessi senza formule chiuse (correlazione, $R^2$, mediana, differenza tra medie/mediane di due gruppi), oppure quando la distribuzione asintotica non è normale.

### Pattern A: bootstrap manuale con loop

Più trasparente dal punto di vista didattico:

```{r boot-manual, eval=FALSE}
set.seed(7369)
B <- 1000
n <- length(x)

theta_boot <- numeric(B)

for (b in 1:B) {
  resample <- sample(x, size = n, replace = TRUE)
  theta_boot[b] <- mean(resample)            # sostituisci con la statistica di interesse
}
```

### Pattern B: con la libreria `bootstrap` (stile prof)

Più compatto, è quello che usa Brusa nelle Solutions:

```{r boot-library, eval=FALSE}
library(bootstrap)

set.seed(7369)
n <- length(x) /nrow se df

# funzione ausiliaria: prende indici, restituisce la statistica
aux_fun <- function(ind) {
  mean(x[ind])                              # sostituire con la statistica di interesse
}

result_boot <- bootstrap::bootstrap(x = 1:n,
                                    nboot = 1000,
                                    theta = aux_fun)

# accesso alle B stime bootstrap
theta_boot <- result_boot$thetastar
```

**Per statistiche su 2 campioni** (es. differenza medie A e B):

```{r boot-two-samples, eval=FALSE}
require(bootstrap)
set.seed(130)
n <- length(serverA)
aux_fun <- function(ind) {
A.boot <- serverA[ind]
B.boot <- serverB[ind]
mean(A.boot) - mean(B.boot)
}
mean.boot <- bootstrap::bootstrap(x = 1:n,
                                  nboot = 1000,
                                  theta = aux_fun)
theta_boot <- mean.boot$thetastar  
summary(theta_boot)
```
```

### Standard error e bias

```{r boot-se-bias, eval=FALSE}
# original point estimate
theta_orig <- mean(x)
round(theta_orig, 3)

# standard error = SD of the B bootstrap estimates
se_boot <- sd(theta_boot)
round(se_boot, 3)

# bias = bootstrap mean - original estimate
bias_boot <- mean(theta_boot) - theta_orig
round(bias_boot, 3)

# SE-to-estimate ratio (useful for commenting)
round(100 * se_boot / theta_orig, 2)  
```

### Intervallo di confidenza con il metodo percentile

```{r boot-ci, eval=FALSE}
ci_95 <- quantile(theta_boot, probs = c(0.025, 0.975))
round(ci_95, 3)

# percentile CI 90%
ci_90 <- quantile(theta_boot, probs = c(0.05, 0.95))
round(ci_90, 3)

# percentile CI 99%
ci_99 <- quantile(theta_boot, probs = c(0.005, 0.995))
round(ci_99, 3)
```

### Plot della distribuzione bootstrap

Pattern standard delle Solutions del prof:

```{r boot-plot, eval=FALSE}
hist(theta_boot,
     breaks = 40,
     freq = FALSE,                           # densità invece di frequenze
     col = "lightgray", border = "white",
     main = "Bootstrap distribution",
     xlab = "Stima bootstrap",
     ylab = "Densità")

# linea verticale alla stima originale
abline(v = theta_orig, col = "red", lwd = 2, lty = 1)

# linea verticale alla media delle B stime bootstrap
abline(v = mean(theta_boot), col = "blue", lwd = 2, lty = 2)

# se hai calcolato il CI: linee tratteggiate ai bound
ci <- quantile(theta_boot, c(0.025, 0.975))
abline(v = ci, col = "green", lwd = 2, lty = 2)

legend("topright",
       legend = c("Stima originale", "Media bootstrap", "CI bounds"),
       col = c("red", "blue", "green"),
       lty = c(1, 2, 2), lwd = 2,
       cex = 0.7)
```

### CI percentile vs CI simmetrico

Il **CI percentile** è preferito al **CI simmetrico** $\hat\theta \pm 1.96 \cdot \widehat{se}$ quando:

- la distribuzione di $\hat\theta$ è **asimmetrica**
- $\hat\theta$ ha **range vincolato** (es. correlazione in $[-1,1]$, varianza $> 0$, proporzione in $[0,1]$)
- la dimensione campionaria è **piccola**

Per la **media** in campioni grandi, i due CI tendono a coincidere (TLC garantisce normalità asintotica).

### Template 1 — Bootstrap of a single statistic (e.g., mean, median, SD, correlation)

**Block A — Setup and motivation**

The accuracy of the estimator $\hat\theta = $ `[describe statistic]` is assessed 
via the nonparametric bootstrap with $B = [B]$ replications. This procedure 
allows estimation of the standard error and the construction of confidence 
intervals without relying on distributional assumptions, which is particularly 
useful given `[reason: small sample size / unknown distribution / non-standard 
estimator]`.

**Block B — Descriptive of the bootstrap distribution**

The $B = [B]$ bootstrap estimates of $\hat\theta$ range from `[min]` to `[max]`, 
with mean `[mean_boot]` and median `[median_boot]`. Both the mean and median 
of the bootstrap distribution are very close to the original sample estimate 
$\hat\theta = [theta_orig]$, in agreement with the general property that the 
bootstrap distribution is centered (in expectation) on the original sample 
estimate.

**Block C — Standard error**

The bootstrap standard error is computed as the standard deviation of the $B$ 
bootstrap estimates: $\widehat{SE}_{\text{boot}} = [SE]$. This represents 
approximately `[SE / |theta_orig| · 100]%` of the magnitude of the original 
estimate, suggesting a `[high / moderate / low]` level of precision. 

minore 15 - alta
15-30 moderata
maggiore 30 alta

[If the sample contains potential outliers, add:] The variability of the 
bootstrap distribution is also influenced by the presence of `[outlier / 
unusual observation identified in Block X]`, which is included in different 
proportions in the various resamples and amplifies the spread of $\hat\theta^*$.

**Block D — Conclusion**

In summary, the bootstrap procedure provides a point estimate of $\hat\theta = 
[theta_orig]$ with standard error of approximately $[SE]$. Confidence intervals 
will be constructed in the next step via the percentile method.

---

### Template 2 — Difference between two means

**Block A — Setup**

To compare `[group_A]` and `[group_B]` in terms of their mean values of 
`[variable]`, the bootstrap is applied to the **difference of sample means** 
$\hat\theta = \bar X_A - \bar X_B$. The sample estimate of this difference 
is $[diff_orig]$ `[units]`, indicating that on average `[group_A]` shows 
`[higher / lower]` values than `[group_B]` by approximately `[|diff_orig|]` 
units.

**Block B — Bootstrap distribution**

With $B = [B]$ bootstrap replications, the differences range from `[min]` to 
`[max]`, with mean `[mean_diff_boot]` (essentially coincident with the 
original difference). The distribution is approximately `[symmetric / 
right-skewed / left-skewed]` around its central value, suggesting that the 
sampling distribution of the difference is `[normal-like / non-normal]`.

**Block C — Standard error and precision**

The bootstrap standard error is $\widehat{SE}_{\text{boot}} = [SE]$ `[units]`. 
This value is `[small / moderate / large]` compared to the magnitude of the 
estimated difference ($\hat\theta / SE = [ratio]$), suggesting that the 
observed difference is `[likely / unlikely]` to be statistically meaningful. 
[A formal confidence interval is constructed in the next step.]

**Block D — Substantive interpretation**

The bootstrap analysis confirms a systematic difference between `[group_A]` 
and `[group_B]` on average, with `[group_A / group_B]` performing better/worse 
by approximately `[|diff|]` units. The small sample size (n = `[n]`) limits 
the precision of this estimate, but the result is consistent with the 
preliminary observations from the descriptive analysis.

---

### Template 3 — Difference between two medians (or other robust statistics)

**Block A — Motivation**

Given the small sample size and the possibility of skewed distributions, the 
bootstrap is also applied to the **difference of medians** as an alternative, 
more robust estimator. The sample estimate is $\hat\theta_{med} = [diff_med_orig]$ 
`[units]`.

**Block B — Bootstrap distribution (key difference from means)**

With $B = [B]$ bootstrap replications, the differences of medians range from 
`[min]` to `[max]`. Unlike the bootstrap distribution of the mean — which is 
typically smooth and continuous — the distribution of the median difference 
takes only a **small number of distinct values**. For example, [observed 
pattern: many replicates give the same value, gaps in the distribution]. 

This discreteness is a known feature of the bootstrap when applied to 
**non-smooth statistics** (such as medians and quantiles) on small samples: 
the bootstrap resampling can only reproduce values already present in the 
original sample, leading to a limited set of possible outcomes for 
quantile-based statistics.

**Block C — Standard error**

The bootstrap standard error of the median difference is `[SE_med]`. This is 
`[higher / lower / similar to]` the SE of the mean difference ($[SE_mean]$), 
which is consistent with the general fact that the median tends to be `[more / 
less]` variable than the mean when `[distribution / sample size]` is 
`[skewed / small / large]`.

**Block D — Implication for confidence intervals**

Due to the discreteness of the bootstrap distribution of the median, the 
percentile confidence interval may be less reliable than for the mean: 
specific quantiles may not correspond to natural cut-off points in the data. 
This will be reflected in the next step, where the percentile interval is 
constructed.

---

### Template 4 — Percentile confidence interval

**Block A — Construction**

The percentile-based confidence interval at the $[1-\alpha]$% confidence level 
is constructed by taking the $\alpha/2$ and $1-\alpha/2$ quantiles of the 
bootstrap distribution:

$$
CI_{[1-\alpha]} = \left[\hat\theta^*_{\alpha/2},\; \hat\theta^*_{1-\alpha/2}\right]
= [L, U]
$$

In our case, the $[1-\alpha]$% confidence interval for `[parameter]` is 
$[L, U]$.

**Block B — Interpretation**

With $[1-\alpha]$% confidence, the true value of `[parameter]` lies in the 
interval $[L, U]$. This means that if we repeated the sampling and bootstrap 
procedure many times, approximately $[1-\alpha]$% of the resulting intervals 
would contain the true value of the parameter.

**Block C — Length of the interval**

The width of the interval is $U - L = [width]$ `[units]`, [representing 
approximately [width/theta_orig · 100]% of the point estimate]. A `[narrow / 
wide]` interval suggests a `[precise / imprecise]` estimate, which reflects 
the [combined effect of the sample size, the variability in the data, and the 
inherent variability of the chosen estimator].

**Block D — Substantive conclusion**

[For a difference of means/medians:]

Since the interval `[does / does not]` include zero, the difference between 
`[group_A]` and `[group_B]` is `[statistically significant / not statistically 
significant]` at the $[α]$% level. Practically, this means that the observed 
difference of $[diff]$ is `[likely / unlikely]` to be due to random sampling 
variability alone.

[For a correlation:]

Since the interval `[does / does not]` include zero, the linear association 
between `[X]` and `[Y]` is `[statistically significant / not statistically 
significant]` at the $[α]$% level. The plausible range of values for the true 
correlation is `[L, U]`, suggesting `[a weak / moderate / strong]` association.

[For a single mean or SD:]

The interval $[L, U]$ provides plausible values for the true population 
parameter, which can be used for substantive interpretation in the application 
context.

---

### Template 5 — Plot of the bootstrap distribution with annotations

### Standard template — bootstrap of a single statistic

The histogram represents the bootstrap distribution of the estimator $\hat\theta = 
[describe statistic]$, obtained from $B = [B]$ replications of resampling with 
replacement from the original sample. The values taken by the bootstrap estimates 
range from a minimum of $[min]$ to a maximum of $[max]$. The distribution is 
approximately [symmetric / asymmetric] around a central peak, located at the 
value of $[mean_boot]$, which coincides with both [the original sample estimate / 
the mean of the $B$ bootstrap estimates]. The spread of the distribution reflects 
the magnitude of the bootstrap standard error, $\widehat{SE}_{\text{boot}} = [SE]$.

### Standard template — bootstrap with annotation of original estimate and bootstrap mean

The histogram of the bootstrap distribution shows the variability of the 
estimator $\hat\theta$ across $B = [B]$ resamples of size $n = [n]$ drawn from 
the original sample. The values range from $[min]$ to $[max]$, with a central 
peak around $[mean_boot]$.

The two vertical lines correspond to the **original sample estimate** ($\hat\theta = 
[theta_orig]$, in [color]) and to the **mean of the $B$ bootstrap replicates** 
($\bar{\hat\theta^*} = [mean_boot]$, in [color]). The two values are practically 
[coincident / very close], in agreement with the expected behaviour of the 
nonparametric bootstrap: the bootstrap distribution is centered, in expectation, 
on the original sample estimate. The width of the histogram around these two 
lines provides a visual representation of the standard error of the estimator.

### Template — bootstrap with confidence interval bounds added

The histogram of the bootstrap distribution is reported below, with vertical lines 
indicating the original sample estimate ($[theta_orig]$, in red), the mean of the 
$B$ bootstrap estimates ($[mean_boot]$, in blue), and the lower and upper bounds 
of the $[1-\alpha]$% percentile confidence interval ($[L]$ and $[U]$, in green, 
dashed).

The shape of the distribution is approximately [symmetric and bell-shaped / 
asymmetric / right-skewed / left-skewed], with the original estimate falling 
[in the center of / near one edge of] the bootstrap distribution. The confidence 
interval [$L$, $U$] covers approximately the central $[1-\alpha]$% of the bootstrap 
mass and represents a range of values centered around the point estimate of the 
parameter.

[Always close with substantive interpretation:]

It is worth noting that $[1-\alpha]\%$ is not the probability that the point 
estimate falls within the interval. Rather, we are confident at the 
$[1-\alpha]\%$ level that the true parameter falls within this range. The 
[narrow / wide] width of the interval reflects [a good / a limited] precision 
in estimating the parameter. No strong asymmetry or multimodality is observed in the bootstrap distribution, supporting the stability of the estimator.

### Template — bootstrap of a discrete statistic (e.g., median on small sample)

The histogram of the bootstrap distribution of the [median difference / median] 
shows a noticeable feature compared to the case of the [mean / smooth statistic]: 
the distribution takes only a **small number of distinct values**, with several 
isolated bars and gaps. This is a known characteristic of the nonparametric 
bootstrap when applied to **non-smooth statistics** — such as medians or 
quantiles — computed on small samples. Since the bootstrap can only reproduce 
values already present in the original sample, the set of possible outcomes for 
quantile-based statistics is limited, and the bootstrap distribution becomes 
discrete rather than smooth.

The two vertical lines, representing the original estimate ($[theta_orig]$) and 
the bootstrap mean ($[mean_boot]$), [are close to each other / show a small 
discrepancy], with both falling near the most frequently observed bootstrap 
value. In such cases, the percentile confidence interval may be wider and more 
discrete than the corresponding interval for the mean.

### Template — bootstrap distribution with outlier influence

The histogram of the bootstrap distribution shows a clear [right / left] 
asymmetry, with a [longer / shorter] [right / left] tail. This asymmetry is 
consistent with the presence of the [outlier / unusual observation] identified 
in the descriptive analysis: bootstrap resamples that include this observation 
multiple times produce estimates that deviate substantially from the central 
mass of the distribution, contributing to the [skewness / heavy tail / 
bimodality].

This pattern justifies the use of the **percentile method** for the construction 
of the confidence interval, as it directly reflects the actual shape of the 
bootstrap distribution. A symmetric (Wald-type) interval, $\hat\theta \pm 1.96 
\cdot \widehat{SE}$, would impose a symmetry not present in the data and could 
yield biased coverage
---

### Template 6 — Comparing percentile CI vs symmetric CI (if relevant)

**Block A — Setup**

The percentile-based bootstrap CI for `[parameter]` is $[L_{perc}, U_{perc}]$, 
while the symmetric Wald-type CI would be $\hat\theta \pm 1.96 \cdot 
\widehat{SE}_{\text{boot}} = [L_{sym}, U_{sym}]$.

**Block B — Comparison**

[If similar:] The two intervals are very close, indicating that the bootstrap 
distribution of $\hat\theta$ is approximately symmetric and roughly normal. 
In this case, the two methods produce comparable inferential conclusions.

[If different:] The two intervals differ noticeably: $[L_{perc}, U_{perc}]$ 
(percentile) vs $[L_{sym}, U_{sym}]$ (symmetric). The percentile interval 
reflects the actual shape of the bootstrap distribution, which is 
`[asymmetric / has heavy tails / shifted]`, while the symmetric interval 
imposes a symmetric form that does not match the data. The percentile 
interval is therefore preferable in this case, as it captures the true 
sampling variability of the estimator more accurately.
### Trappole comuni

- **Resampling senza ripetizione** è inutile: produrrebbe solo permutazioni del campione originale, $\hat\theta$ non cambierebbe (è simmetrico nei dati)
- Il **`set.seed()`** è obbligatorio in ogni chunk di bootstrap, altrimenti non riproducibile
- **Bootstrap della mediana / quantili**: con $n$ piccolo, la distribuzione bootstrap è discreta e ha pochi valori distinti — i CI possono essere imprecisi (vedi Es. 12 nelle Solutions)
- Il bootstrap **non corregge** errori sistematici: se il campione originale è distorto, anche il bootstrap lo è

# Linear Regression — Cheat Sheet (Chapters 4–10)

> Nota: tutti gli snippet sono testati e basati sul codice ufficiale del corso (Applications + Solutions Brusa). Pronti da copiare e adattare.

---

## Model fit and reading the summary

### Specification

Multiple linear regression model:

$$Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_i$$

**Five assumptions** (always state them in your analysis):

1. **Linearity**: response is a linear and additive function of covariates
2. **Zero-mean errors**: $\mathbb{E}[\epsilon_i] = 0$
3. **Homoscedasticity**: $\text{Var}(\epsilon_i) = \sigma^2$ constant for all $i$
4. **Independence**: $\text{Cov}(\epsilon_i, \epsilon_j) = 0$ for $i \neq j$
5. **Normality** (needed for t-test, F-test, CI): $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$

### Research question
The main objective is to model how [RESPONSE] depends on [COVARIATES],
in the context of [APPLIED CONTEXT].

Specifically, we aim to:
1. Quantify the effect of [COVARIATES] on [RESPONSE]
2. Identify which covariates are significant predictors
3. [Optional: predict / make decisions / classify / ...]

Based on [theory / domain knowledge], we expect [HYPOTHESES ABOUT SIGNS].
The analysis will [verify these hypotheses / explore the relationships].

The available data come from an observational study, not from a randomized experiment. Therefore, the estimated effect of [explanatory variable] on [response variable] should be interpreted as an association rather than a causal relationship. In particular, unmeasured confounding factors or external influences may partially explain the observed relationship. (observed and not recorded with any kind of experiment)

### Fitting the model

```{r lm-fit, eval=FALSE}
# fit a multiple linear regression
mod <- lm(y ~ x1 + x2 + x3, data = df)

# fit using all available covariates
mod_full <- lm(y ~ ., data = df)

# print results
summary(mod)
```

### Reading the summary

The output of `summary(lm)` is structured as follows:

```
Coefficients:
            Estimate Std. Error t value Pr(>|t|)    
(Intercept)  12.345    1.234     10.0   <2e-16 ***  ← intercept
x1            0.567    0.123      4.6   1e-05  ***  ← coefficient for x1
x2           -0.234    0.089     -2.6   0.012  *    ← coefficient for x2

Residual standard error: 0.95 on 47 degrees of freedom   ← S, sqrt(RSS/(n-p-1))
Multiple R-squared:  0.78,    Adjusted R-squared:  0.76    ← R^2 and R^2-adj
F-statistic: 55.2 on 3 and 47 DF,  p-value: < 2.2e-16    ← global F-test
```

**What to comment**:

- **Coefficients**: estimated value, sign (matches expectation?), magnitude, significance (`***`, `**`, `*`, `.`)
- **Residual standard error**: estimate of $\sigma$. Smaller is better
- **R-squared**: fraction of variance explained
- **Adjusted R-squared**: same but penalized for number of parameters (use it to compare models with different $p$)
- **F-statistic**: tests $H_0$: all $\beta_k = 0$. If significant, at least one covariate is useful

### Extracting components

```{r lm-extract, eval=FALSE}
coef(mod)              # estimated coefficients
fitted(mod)            # predicted values y_hat
residuals(mod)         # residuals (y - y_hat)
confint(mod)           # 95% CI for coefficients
confint(mod, level = 0.99)   # custom confidence level
sigma(mod)             # residual standard error
summary(mod)$r.squared       # R^2
summary(mod)$adj.r.squared   # adjusted R^2
```

### Estimated model

The estimated model is:

$\hat{Y} = [β₀] + [β₁] \cdot \text{[var1]} + [β₂] \cdot \text{[var2]} + \dots$

The intercept ($[β₀]$) [does/does not] have a meaningful interpretation, since 
[explain]. As expected from our research question, the slope coefficients are 
[brief description], consistent with the hypothesis that [substantive 
interpretation].

### Coefficient interpretation

- **[var1]** ($\hat{\beta}_1 = [value]$): [significant/not significant] at the 
  [α]% level (p-value $[p]$). A unit increase is associated with an average 
  [change] of $[β_k]$ units in $[response]$, ceteris paribus.

- **[var2]** ($\hat{\beta}_2 = [value]$): [...]

- **[var3]** ($\hat{\beta}_3 = [value]$): [if non-significant: discussion of 
  why this might be — multicollinearity, sample size, etc.]

The estimated standard errors of the regression coefficients are reported 
in the second column of the `summary()` output. They quantify the **precision** 
of each coefficient estimate: smaller standard errors correspond to more 
precise estimates.

The intercept has standard error `[SE_intercept]`, while the slope coefficients 
have standard errors ranging from `[SE_min]` to `[SE_max]`. The covariates 
with the `[smallest / largest]` standard error are `[cov_names]`, indicating 
that their effects are estimated with the highest precision. Conversely, 
`[other covariates]` have larger standard errors, reflecting greater 
uncertainty in their estimates.

### Goodness of fit and global test

The model explains $[R²×100]\%$ of the variance in $[response]$ 
($R^2 = [R²]$, adjusted $R^2 = [R²-adj]$). The residual standard error is 
$[S]$, it means that, on average, the observed
and the estimated values for each sample unit differ by [],  Although the value seems small, it needs to be further evaluated with respect to a centrality index (e.g., the mean) of the response variable. Computing the ratio between the RSE and the mean of the carats, we
obtain the percentage error, equal to [RSE/MEAN] = []: on average the percentage errorbetween observed and fitted values is equal to %.

The global F-test yields $F = [F]$ on $[p]$ and $[n-p-1]$ degrees of freedom 
(p-value $[value]$). We reject $H_0$: at least one covariate is associated 
with $[response]$ at this confidence level.

### Preliminary residual analysis

The [raw / deviance] residuals range from `[Min]` to `[Max]`, with median 
`[Median]` — `[close to / not close to]` zero, as expected for a 
[regression / logistic] model with intercept. The interquartile range 
`[Q1, Q3]` is `[approximately centered on / shifted away from]` zero, 
indicating `[symmetry / asymmetry]` in the bulk of the distribution.

[Pick the variant matching what you observe:]

**Variant A — symmetric residuals:** The residuals appear approximately 
symmetric around zero, with $|Min| \approx |Max|$ and $|Q1| \approx |Q3|$. 
The ratio between total range and IQR is approximately `[range/IQR]`, 
consistent with a roughly normal-like distribution. No major concerns 
emerge at this preliminary stage.

**Variant B — right-skewed (Max >> |Min|):** The maximum residual (`[Max]`) 
is noticeably larger in absolute value than the minimum (`[Min]`), 
indicating a slight positive asymmetry in the tails. This may suggest one or 
more observations that the model `[underpredicts / fails to capture]`, 
which warrants verification in the diagnostic plots and Cook's distance.

**Variant C — left-skewed (|Min| >> Max):** The minimum residual is larger 
in absolute value than the maximum, indicating a slight negative asymmetry 
in the tails. This pattern will be further examined via the Q-Q plot to 
assess whether it is a substantial deviation from normality or attributable 
to a single extreme observation.

### One-line tail closure (optional, useful)

For a full assessment of the assumptions, refer to the diagnostic plots 
(`plot(mod)`) and the analysis of leverage / Cook's distance — preliminary 
indications from the residual summary will be confirmed (or revised) in 
that step.


### Looking at the results in light of our initial research question:

- We hypothesized that [hypothesis 1, e.g., "all covariates would be positively 
  associated with Y"]: this is [confirmed / partially confirmed / not 
  confirmed]. Specifically, [variable A] and [variable B] show the expected 
  positive association, while [variable C] [does not / shows the opposite].

- The most influential covariates appear to be [list], based on the magnitude 
  and significance of their coefficients.

- A non-trivial finding is that [unexpected result], which suggests [possible 
  interpretation or follow-up investigation].

---

## Diagnostics: residuals, leverage, Cook, VIF

### Residuals vs continuous covariates

```{r residuals-vs-covariates-base, eval=FALSE}
res <- residuals(mod_final)        # deviance residuals (default for glm)

par(mfrow = c(1, 2))

plot(df$cov1, res,
     main = "Residuals vs cov1",
     xlab = "cov1 label",
     ylab = "Deviance residuals",
     pch = 16, col = "blue")
abline(h = 0, lty = 2, col = "gray")

plot(df$cov2, res,
     main = "Residuals vs cov2",
     xlab = "cov2 label",
     ylab = "Deviance residuals",
     pch = 16, col = "blue")
abline(h = 0, lty = 2, col = "gray")

par(mfrow = c(1, 1))
```

### What to look for

| Pattern | What it suggests |
|---------|------------------|
| Random cloud around 0 | Linearity OK, no concerns |
| Curved (U-shape, parabola) | Non-linear relationship — add quadratic term or transform |
| Monotonic trend | Model systematically over/under-predicts at certain covariate values |
| Funnel shape (variance grows with covariate) | Heteroscedasticity |
| Isolated extreme points | Potential outliers — check Cook's distance |

### Note for logistic regression

In `glm` with binary response, deviance residuals often form **two horizontal 
bands** — one above zero (for observations with $y = 1$) and one below 
(for $y = 0$). This is expected and **not** a violation of model assumptions. 
What you look for is whether each band shows systematic trends or whether 
specific points lie far from both bands.

### Comment template

The two scatterplots show the residuals of the selected model against the 
continuous covariates `[cov1]` and `[cov2]`.

**Residuals vs `[cov1]`**: The residuals are [scattered randomly / show a 
clear pattern] around zero, with [no apparent / a slight / a clear] trend 
across the range of `[cov1]`. The dispersion appears [approximately constant 
/ to increase / to decrease] with `[cov1]`. [No specific concern / A pattern 
that warrants investigation] emerges from this plot.

**Residuals vs `[cov2]`**: The residuals against `[cov2]` show [a similar 
random pattern / a non-linear shape / clear heteroscedasticity]. 
[Specific observation: e.g., "a small group of observations with high `[cov2]` 
shows larger positive residuals, suggesting that the model underpredicts the 
response for these units"].

Overall, the residual plots [support / call into question] the assumption 
of linearity of the [response / logit] in the continuous covariates. 
[Optional: "Possible remedies include adding a quadratic term, a log 
transformation of the covariate, or using a more flexible model."]

### The four diagnostic plots

```{r diagnostic-plots, eval=FALSE}
par(mfrow = c(2, 2))
plot(mod)
par(mfrow = c(1, 1))
```

R produces four plots — what to look for in each:

| Plot | Pattern OK | Pattern problematic | Assumption violated |
|------|------------|---------------------|---------------------|
| Residuals vs Fitted | Random cloud around 0 | U-shape / parabola | Linearity |
| Residuals vs Fitted | Random cloud around 0 | Funnel shape | Homoscedasticity |
| Normal Q-Q | Points on line | Points deviating in tails | Normality |
| Scale-Location | Horizontal red line | Tilted/curved red line | Homoscedasticity |
| Residuals vs Leverage | No points beyond Cook contours | Points outside dashed lines | Influential observations |

### Detecting unusual observations

```{r unusual-observations, eval=FALSE}
# leverage values (hat values)
h <- hatvalues(mod)
n <- nrow(df)
p <- length(coef(mod)) - 1
threshold_lev <- 2 * (p + 1) / n
which(h > threshold_lev)              # high-leverage observations

# standardized residuals
rstd <- rstandard(mod)
which(abs(rstd) > 2)                  # potential outliers (residuals)

# Cook's distance
cook <- cooks.distance(mod)
which(cook > 1)                        # influential observations (strict)
which(cook > 4/n)                      # influential (more sensitive threshold)

# combined diagnostic plot
plot(cook, type = "h", main = "Cook's distance",
     xlab = "Observation index", ylab = "Cook's D")
abline(h = 4/n, lty = 2, col = "red")
```

### Diagnostic summary table

| Type | What's anomalous | Measure | Indicative threshold |
|------|------------------|---------|---------------------|
| Outlier | Residual $z_i$ | Standardized residuals | $\|r^*\| > 2$ or $> 3$ |
| Leverage point | Covariates $x_i$ | Hat values $h_{ii}$ | $> 2(p+1)/n$ |
| Influential point | Effect on $\hat\beta$ | Cook's distance $D_i$ | $> 1$ (strict), $> 4/n$ (sensitive) |

> **Key distinction**: high leverage does NOT imply influential. A point with $h_{ii} = 0.4$ and $r^* \approx 0$ is on the leverage edge but lies on the regression line, so removing it doesn't change $\hat\beta$ → not influential.

### Multicollinearity: VIF

```{r vif, eval=FALSE}
library(car)

vif(mod)                    # VIF for each covariate
```

**Interpretation thresholds**:

| VIF | Diagnosis |
|-----|-----------|
| ~1 | No multicollinearity |
| 1–5 | Tolerable |
| 5–10 | Worth attention |
| > 10 | Serious problem |

**Symptoms of multicollinearity**:

- F-test globally significant but no individual t-test significant
- Coefficient signs unexpected (negative when theory says positive)
- Standard errors very large relative to coefficient magnitudes
- High correlations in `cor(df_covariates)`

**Remedies if VIF too high**:

1. Remove one of the correlated variables (the less interpretable / less theoretically grounded)
2. Combine them into a single derived variable (sum, average, principal component)
3. Increase sample size (reduces SE but not collinearity itself)
4. Use ridge regression (out of scope for this course)

### Comment templates

For diagnostics:

> The Residuals vs Fitted plot shows `[a random cloud around zero / a U-shaped pattern / a funnel shape]`, suggesting that the assumption of `[linearity / homoscedasticity]` `[is / is not]` satisfied. The Normal Q-Q plot indicates `[approximate normality / heavy tails / asymmetry]`.

For unusual observations:

> Observation `[ID]` has a Cook's distance of `[value]`, exceeding the threshold of `[4/n or 1]`. It also shows `[high leverage / a large standardized residual]`. This observation has a substantial influence on the parameter estimates and warrants further investigation. Removing it changes the estimated coefficient of `[variable]` from `[old value]` to `[new value]`.

For multicollinearity:

> The variable `[name]` has a VIF of `[value]`, exceeding the threshold of 10 (or 5). This indicates substantial multicollinearity, which inflates the standard errors of the regression coefficients and reduces the power of the t-tests. The high VIF is consistent with the strong correlation observed between `[variable]` and `[other variable]` in the correlation matrix.

---

## Inference: F-test, t-test, confidence intervals

### Global F-test

Tests $H_0$: all coefficients (except intercept) are zero. Already shown in the bottom of `summary(mod)`. To extract:

```{r f-test, eval=FALSE}
f_stat <- summary(mod)$fstatistic
f_stat                     # value, df1, df2

# p-value of the F-test
pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = FALSE)
```

### Nested F-test (compare two models)

Tests whether a subset of coefficients can all be set to zero (i.e., comparing a reduced model to a full one).

```{r nested-f, eval=FALSE}
# fit reduced and full models
mod_reduced <- lm(y ~ x1, data = df)
mod_full    <- lm(y ~ x1 + x2 + x3, data = df)

# F-test for nested comparison
anova(mod_reduced, mod_full)
```

The output gives the F-statistic, df, and p-value. If significant → the additional covariates are jointly useful.

### Individual t-tests

Already shown in `summary(mod)` (column `Pr(>|t|)`). Tests $H_0: \beta_k = 0$ for each $k$ separately.

**Quick rule of thumb**: $|t| > 2$ implies significance at the 5% level (for $n$ reasonably large).

### Confidence intervals for coefficients

```{r confint, eval=FALSE}
confint(mod)                       # 95% CI for all coefficients
confint(mod, level = 0.99)         # 99% CI
confint(mod, parm = "x1")          # CI for a specific coefficient
```

**Manual calculation** (useful for understanding):

```{r confint-manual, eval=FALSE}
# CI 95% for beta_k
beta_hat <- coef(mod)["x1"]
se_beta  <- sqrt(diag(vcov(mod)))["x1"]
t_critical <- qt(0.975, df = mod$df.residual)

ci_lower <- beta_hat - t_critical * se_beta
ci_upper <- beta_hat + t_critical * se_beta
c(ci_lower, ci_upper)
```

### Comment templates

For F-test:

> The global F-test yields F = `[value]` with `[df1]` and `[df2]` degrees of freedom and p-value `[value]`. We `[reject / fail to reject]` $H_0$ at the 5% level: the model `[provides a significantly better fit than the null model / is not significantly better than the null model]`.

For t-test:

> The coefficient of `[variable]` is `[value]` with t-value `[value]` and p-value `[value]`. We `[reject / fail to reject]` $H_0: \beta_k = 0$ at the `[α]` level. Hence, `[variable]` `[is / is not]` significantly associated with the response, holding the other covariates constant.

For CI:

> The 95% confidence interval for $\beta_{[variable]}$ is `[lower, upper]`. We are 95% confident that the true coefficient lies in this interval. Since the interval `[contains / does not contain]` zero, the coefficient is `[not / yes]` significantly different from zero at the 5% level.

---

## Model selection: AIC, BIC, stepwise

### Information criteria

```{r aic-bic, eval=FALSE}
AIC(mod)              # Akaike Information Criterion
BIC(mod)              # Bayesian Information Criterion (more conservative)

# compare multiple models
AIC(mod1, mod2, mod3)
BIC(mod1, mod2, mod3)
```

**Lower is better.** Always compare across models on the **same dataset**.

### Stepwise selection

```{r stepwise, eval=FALSE}
library(MASS)

# starting from the full model, remove covariates (backward)
mod_back <- stepAIC(mod_full, direction = "backward", trace = TRUE)

# starting from the null model, add covariates (forward)
mod_null <- lm(y ~ 1, data = df)
mod_fwd  <- stepAIC(mod_null,
                    scope = list(lower = ~ 1,
                                 upper = ~ x1 + x2 + x3 + x4),
                    direction = "forward")

# both directions (recommended default)
mod_step <- stepAIC(mod_full, direction = "both", trace = TRUE)

# using BIC instead of AIC: change penalty
n <- nrow(df)
mod_step_bic <- stepAIC(mod_full, direction = "both", k = log(n))

# final selected model
summary(mod_step)
```

The `trace = TRUE` option prints the progress at each step: which variable is added/removed and the resulting AIC.

### Reading the stepwise output

```
Start: AIC=42.5
y ~ x1 + x2 + x3 + x4

       Df  Sum of Sq   RSS    AIC
- x4    1     0.012   12.5   40.5    ← removing x4 lowers AIC
<none>             12.4   42.5
- x3    1     0.5     12.9   44.0    ← removing x3 raises AIC
- x1    1     8.0     20.4   52.0    ← x1 most important

Step:  AIC=40.5
y ~ x1 + x2 + x3
...
```

The `<none>` row indicates the current model. If a `-x` row has lower AIC than `<none>`, that variable will be removed.

### Comment templates

For stepwise selection:

> Stepwise selection based on AIC was performed starting from the full model with `[p]` covariates. The procedure `[removed / retained]` `[variable names]` at each step. The final selected model includes `[selected variables]` and has AIC = `[value]`, lower than the full model (AIC = `[value]`).

For model comparison:

> Comparing AIC and BIC: AIC selects a model with `[k1]` covariates, while BIC selects a more parsimonious model with `[k2]` covariates. Given that BIC penalizes complexity more heavily ($\log(n) = [value] > 2$ for our sample size), this difference is expected. `[Choice based on objective: AIC for prediction, BIC for interpretability.]`

---

## Prediction: confidence interval vs prediction interval

### Two types of prediction

| Goal | Question | Interval type | Width |
|------|----------|---------------|-------|
| Mean response | "What is the average $Y$ for units with $x_c$?" | Confidence interval (CI) | Narrower |
| Single new observation | "What is $Y$ for THIS new unit with $x_c$?" | Prediction interval (PI) | Wider |

The PI is wider because it includes the irreducible variance $\sigma^2$ of the individual error term, in addition to the uncertainty about $\hat\beta$.

### Computation in R

```{r prediction, eval=FALSE}
# new observation with specified covariate values
new_data <- data.frame(x1 = 5, x2 = 2.3, x3 = "B")

# point prediction
predict(mod, newdata = new_data)

# 95% CI for the mean response (confidence interval)
predict(mod, newdata = new_data, interval = "confidence", level = 0.95)

# 95% PI for a single new observation (prediction interval)
predict(mod, newdata = new_data, interval = "prediction", level = 0.95)
```

Output structure:

```
       fit     lwr      upr
1   23.45   21.89   25.01    ← CI: tighter
1   23.45   18.32   28.58    ← PI: wider
```

### Plotting prediction bands

```{r predict-plot, eval=FALSE}
# range of x for plotting
x_range <- seq(min(df$x1), max(df$x1), length.out = 100)
new_data <- data.frame(x1 = x_range)

# both intervals
pred_ci <- predict(mod_simple, newdata = new_data, interval = "confidence")
pred_pi <- predict(mod_simple, newdata = new_data, interval = "prediction")

plot(df$x1, df$y, pch = 16, col = "blue",
     main = "CI vs PI", xlab = "x1", ylab = "y")
lines(x_range, pred_ci[, "fit"], lwd = 2, col = "red")
lines(x_range, pred_ci[, "lwr"], lty = 2, col = "red")
lines(x_range, pred_ci[, "upr"], lty = 2, col = "red")
lines(x_range, pred_pi[, "lwr"], lty = 2, col = "darkgreen")
lines(x_range, pred_pi[, "upr"], lty = 2, col = "darkgreen")
legend("topleft",
       legend = c("Fit", "CI 95%", "PI 95%"),
       col = c("red", "red", "darkgreen"),
       lty = c(1, 2, 2), lwd = 2, cex = 0.8)
```

### Comment template

> For a new unit with covariates `[values]`, the predicted response is `[value]`. The 95% confidence interval for the mean response is `[L_ci, U_ci]`: with 95% confidence, the average response for units with these covariate values lies in this interval. The 95% prediction interval for a single new observation is `[L_pi, U_pi]`, wider because it accounts for the irreducible individual variability $\sigma^2$ in addition to the uncertainty in the estimated coefficients.

---

## Categorical covariates and interactions

### Dummy coding (automatic in R)

R automatically creates $J-1$ dummies for a categorical variable with $J$ levels. The first level (alphabetical order) becomes the reference.

```{r factors-basics, eval=FALSE}
# verify factor structure
str(df$category)
levels(df$category)        # first level is the reference

# convert character to factor
df$category <- as.factor(df$category)

# change reference level
df$category <- relevel(df$category, ref = "specific_level")

# fit a model with categorical covariate
mod <- lm(y ~ x_continuous + x_categorical, data = df)
summary(mod)
```

### Interpretation of coefficients

For a categorical variable with levels A (reference), B, C, D:

```
Coefficients:
              Estimate    
(Intercept)    25000        ← E[Y] for reference category A (other covariates = 0)
xCategoryB     1500         ← E[Y in B] - E[Y in A], holding others constant
xCategoryC    -800
xCategoryD     2100
```

### Parallel-lines model (no interaction)

```{r parallel-lines, eval=FALSE}
mod_parallel <- lm(y ~ x1 + group, data = df)
```

Same slope $\beta_1$ for all groups, different intercepts.

### Model with interaction

```{r interaction, eval=FALSE}
# x1 * group expands to x1 + group + x1:group
mod_interact <- lm(y ~ x1 * group, data = df)
summary(mod_interact)

# only the interaction term, without main effects (rare):
mod_inter_only <- lm(y ~ x1:group, data = df)
```

For binary `group` (0/1):

| Group | Slope of $x_1$ | Intercept |
|-------|----------------|-----------|
| 0 (reference) | $\beta_1$ | $\beta_0$ |
| 1 | $\beta_1 + \gamma$ | $\beta_0 + \delta$ |

The coefficient $\gamma$ on the interaction term measures **how much the slope changes** between groups.

### Testing the interaction

```{r interaction-test, eval=FALSE}
# t-test on the interaction coefficient (in summary)
summary(mod_interact)

# nested F-test: with vs without interaction
anova(mod_parallel, mod_interact)
```

If the F-test or t-test on the interaction is significant → the interaction is needed (groups have different slopes). Otherwise → the parallel-lines model is preferable for parsimony.

### Visualizing interactions

```{r interaction-plot, eval=FALSE}
# scatterplot with separate fitted lines per group
plot(df$x1, df$y,
     col = as.numeric(df$group),
     pch = 16,
     main = "Interaction visualization",
     xlab = "x1", ylab = "y")

# add fitted lines for each group
abline(intercept, beta2, col = "brown")
abline(intercept + beta1, beta2, col = "blue")
```


```{r, eval=FALSE}
# scatterplot with non-parallel fitted lines (different slopes per group)
library(ggplot2)

ggplot(df, aes(x = VAR_CONTINUA, y = RESPONSE, color = VAR_GRUPPO)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("salmon", "steelblue")) +
  labs(title = "Interaction model: RESPONSE vs VAR_CONTINUA",
       x = "VAR_CONTINUA label",
       y = "RESPONSE label",
       color = "VAR_GRUPPO label") +
  theme_bw()
```

## Comment templates — Categorical covariates and interactions

### Template 1 — Categorical covariate (with one or more levels)

**Block 1 — Introduction and significance statement**

The variable `[category]` is included in the model as a categorical predictor 
with `[J]` levels: `[list levels]`. R automatically encodes it through `[J-1]` 
dummy variables, taking `[ref]` as the reference category. The estimated 
coefficients of the dummies represent the **average difference in the response** 
between each non-reference level and `[ref]`, holding all other covariates 
constant.

**Block 2 — Coefficient-by-coefficient interpretation**

The coefficient associated with category `[B]` is $\hat\beta_{[B]} = [value]$, 
with standard error $[SE]$ and `[t / z]`-statistic $[value]$ (p-value $[value]$). 
This indicates that, compared to the reference category `[ref]`, units in `[B]` 
have on average a `[higher / lower]` value of the response by approximately 
$[|value|]$ units, holding the other covariates constant. The effect is 
**[statistically significant / not statistically significant]** at the 
$[α]$% level.

[Repeat the paragraph above for each non-reference level.]

**Block 3 — Substantive interpretation**

In substantive terms, this means that `[contextual interpretation: e.g., 
"insulated houses consume on average 1.57 cubic feet less gas per week than 
non-insulated ones, at any given temperature"]`. This result is `[consistent 
with / contrary to]` our initial hypothesis that `[recall hypothesis from 
research question]`.

**Block 4 — Robustness note (optional but appreciated)**

Note that changing the reference category would alter the numerical values of 
the dummy coefficients but would **not** affect: the fitted values $\hat{y}_i$, 
the residual standard error, the $R^2$, the global F-test, or any predictions. 
This is a general property of dummy coding: the model is invariant to the 
choice of reference; only the parameterization changes.


### Template 2 — Parallel-lines vs interaction model

**Block 1 — Setup of the comparison**

To assess whether the effect of `[continuous_covariate]` on the response 
differs across the levels of `[categorical_covariate]`, two competing 
specifications are considered:

- **Parallel-lines model** (no interaction): assumes that the slope of 
  `[continuous_covariate]` is identical across groups defined by 
  `[categorical_covariate]`. The groups differ only in their intercept 
  (a constant additive shift).
  
$$
Y_i = \beta_0 + \beta_1 \cdot \text{[x1]}_i + \delta \cdot D_i + \epsilon_i
$$

- **Interaction model**: allows the slope of `[continuous_covariate]` to 
  vary across groups, in addition to a possible difference in intercept.

$$
Y_i = \beta_0 + \beta_1 \cdot \text{[x1]}_i + \delta \cdot D_i + \gamma \cdot (\text{[x1]}_i \cdot D_i) + \epsilon_i
$$

The two models are nested: the parallel-lines model corresponds to imposing 
$\gamma = 0$ in the interaction model.

**Block 2 — Formal test**

A **nested F-test** compares the two specifications:

```{r f-test1, eval=FALSE}
anova(mod_parallel, mod_interaction)
```

The test statistic is $F = [value]$ on $(1, [n-p-1])$ degrees of freedom, 
with p-value $[value]$. We **[reject / fail to reject]** the null hypothesis 
$H_0: \gamma = 0$ (i.e., that the slopes are equal across groups) at the 
$[α]$% level. The same conclusion can be drawn by inspecting the t-test on 
the interaction coefficient $\hat\gamma$ in the summary of the interaction 
model, which yields t = $[value]$ and p-value $[value]$.

**Block 3 — Choice of model**

Based on this result, the **[parallel-lines / interaction]** model is 
preferred:

- If parallel-lines is preferred: the data do not provide enough evidence 
  that the slopes differ across groups, so the more parsimonious parallel-lines 
  specification is retained. This is consistent with the principle of 
  preferring simpler models when the additional complexity does not yield 
  significant improvements in fit.
- If interaction is preferred: the slopes are significantly different across 
  groups, and forcing them to be equal would misspecify the model.

**Block 4 — Interpretation of slopes (use this only if interaction is preferred)**

Under the selected interaction model:

- The estimated slope of `[continuous_covariate]` for the reference group 
  `[A]` is $\hat\beta_1 = [value1]$.
- The estimated slope for group `[B]` is $\hat\beta_1 + \hat\gamma = 
  [value1 + value_gamma] = [combined_value]$.
- The difference $\hat\gamma = [value_gamma]$ measures **how much the slope 
  changes** when moving from group `[A]` to group `[B]`. In substantive terms, 
  this means that the effect of `[continuous_covariate]` on the response is 
  **[stronger / weaker]** in group `[B]` than in group `[A]`.

**Block 5 — Interpretation of intercepts (always include)**

The intercept differs between the two groups by the amount $\hat\delta = 
[value]$: at the reference value of `[continuous_covariate]` (typically zero 
or its sample mean if centered), units in group `[B]` have an average 
response that is `[higher / lower]` than units in group `[A]` by $[|value|]$ 
units.

**Block 6 — Summary and transition**

Overall, `[brief summary: e.g., "the two groups differ both in baseline level 
and (if interaction) in their sensitivity to the continuous covariate"]`. 
This finding has practical implications for `[contextual application]`. The 
next analytical step is to `[fit diagnostics / predict for new units / 
compute confidence intervals for the slopes]`.

ù
### Quick reference table

| Scenario | Test result | Preferred model | Substantive interpretation |
|----------|-------------|------------------|----------------------------|
| F-test p > 0.05 | Fail to reject $H_0: \gamma = 0$ | Parallel-lines | Effect of $x_1$ is the same across groups; groups differ only in baseline |
| F-test p < 0.05 | Reject $H_0: \gamma = 0$ | Interaction | Effect of $x_1$ depends on the group; modify wording of interpretation |
| F-test p ≈ 0.05 | Borderline | Either, prefer simpler unless theory supports interaction | Report both and let theory guide |

### Practical tip

In R, the nested F-test for parallel-lines vs interaction is performed automatically when you write:

```{r nested-f1, eval=FALSE}
mod_parallel    <- lm(y ~ x1 + group, data = df)
mod_interaction <- lm(y ~ x1 * group, data = df)    # x1 * group = x1 + group + x1:group

anova(mod_parallel, mod_interaction)
```

The output will show the F-statistic, df, and p-value of the test.

---

### Concrete fill-in for the whiteside (Es. 23) example

Adapt the template to your specific exercise as follows:

The variable `Insul` enters the model as a categorical predictor with two 
levels — `Before` (reference) and `After`. R encodes it as a single dummy 
$D_i$ equal to $1$ for after-insulation observations and $0$ for before. 
The estimated coefficient $\hat\beta_{InsulAfter} = -1.57$ (with SE $= 
0.20$, t = $-7.83$, p-value $< 0.001$) indicates that, holding the external 
temperature constant, after-insulation observations show on average a gas 
consumption $1.57$ cubic feet **lower** than before-insulation observations. 
The effect is highly statistically significant.

In substantive terms, this means that the insulation intervention is 
associated with a reduction of approximately $1.57$ cubic feet of weekly 
gas consumption, at any given outdoor temperature. The result is consistent 
with our initial hypothesis that thermal insulation reduces heating demand.

[If you were to test for interaction:]

A nested F-test was conducted to assess whether the slope of `Temp` differs 
between the two insulation periods. The F-statistic is $[value]$ on $(1, 51)$ 
degrees of freedom (p-value $[value]$): we [reject / fail to reject] the 
hypothesis of equal slopes. The [parallel-lines / interaction] specification 
is therefore preferred. [Continue with the appropriate interpretation block.]

## Common pitfalls (regression)

- **R² always increases** when adding covariates, even useless ones → use adjusted R² for model comparison
- **Significant F + non-significant t-tests** → typical sign of multicollinearity (check VIF)
- **High leverage ≠ influential** → a leverage point with small residual is on the regression line and doesn't change $\hat\beta$
- **Statistical significance ≠ practical importance** → a coefficient with p < 0.001 may have a tiny effect size; always report magnitude alongside p-values
- **Model selection ≠ causal identification** → AIC/BIC find good predictive models, not causal effects
- **In GLMs**: the t-test in `summary` is a Wald test; for nested comparisons use `anova(mod, test = "Chisq")` (likelihood ratio test)
- **Categorical variable read as numeric** → wrong model; check with `str(df)` and convert with `as.factor()` if needed


# Regressione Logistica

## Quando usarla

Usa la regressione logistica quando la **risposta è binaria** ($Y \in \{0, 1\}$) e vuoi modellare la **probabilità di successo** in funzione di covariate (continue, categoriali, o entrambe).

**Perché non `lm()` su una risposta binaria?**

- Le predizioni possono uscire da $[0, 1]$
- Gli errori non sono normali (seguono una Bernoulli)
- La varianza non è costante: $\text{Var}(Y_i) = p_i(1 - p_i)$ dipende da $p_i$ (eteroschedasticità intrinseca)

## La trasformazione logit

| Quantità | Formula | Range |
|----------|---------|-------|
| Probabilità di successo | $p_i = P(Y_i = 1)$ | $[0, 1]$ |
| Odds | $O_i = p_i / (1 - p_i)$ | $[0, +\infty)$ |
| Logit | $\log(O_i) = \log[p_i / (1 - p_i)]$ | $(-\infty, +\infty)$ |

Il logit viene modellato linearmente:

$$\text{logit}(p_i) = \log\left(\frac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip}$$

Trasformazione inversa (per tornare alla probabilità):

$$p_i = \frac{\exp(\beta_0 + \beta_1 x_{i1} + \dots)}{1 + \exp(\beta_0 + \beta_1 x_{i1} + \dots)}$$

Questa è la **funzione logistica** — una curva a S da 0 a 1.

## Workflow checklist

Per qualsiasi esercizio di regressione logistica, segui questo ordine:

1. **Setup**: caricamento dati, gestione NA, conversione della risposta a factor
2. **EDA**: descrittive per livello di risposta, contingency tables o boxplot
3. **What are the logits**: rispondi (se richiesto) o tienilo a mente
4. **Fit**: `glm(y ~ x1 + x2 + ..., family = binomial, data = df)`
5. **Lettura summary**: Wald test, null/residual deviance, AIC
6. **Selezione**: `stepAIC()` se richiesto
7. **Interpret coefficients**: odds ratio $\exp(\hat\beta_k)$ con CI
8. **Predict**: probabilità per nuove unità con `type = "response"`
9. **Compare nested models** (opzionale): `anova(M0, M1, test = "Chisq")`

---

## Step 1 — Setup

```{r logit-setup, eval=FALSE}
# remove rows with NA on covariates of interest
df_clean <- df[complete.cases(df[, c("y", "x1", "x2")]), ]
nrow(df_clean)

# response as factor with proper levels
df_clean$y <- as.factor(df_clean$y)
levels(df_clean$y)                  # first level = reference (= "failure" / 0)

# if needed, change the reference level
df_clean$y <- relevel(df_clean$y, ref = "no")   # "no" becomes reference

# verify
table(df_clean$y)
```

---

## Step 2 — Exploratory Analysis

### Case A: Continuous covariates

```{r eda-continuous, eval=FALSE}
require(gridExtra)
require(ggplot2)

p1 <- ggplot(df, aes(x = response, y = covariate1)) +
  geom_boxplot(aes(fill = response), col = "black") +
  geom_point(aes(col = response)) +
  xlab("[response label]") + ylab("[covariate1 label]") +
  scale_fill_manual(values = c("#ffe5ab", "#cce8f8")) +
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  theme_bw()

p2 <- ggplot(df, aes(x = response, y = covariate2)) +
  geom_boxplot(aes(fill = response), col = "black") +
  geom_point(aes(col = response)) +
  xlab("[response label]") + ylab("[covariate2 label]") +
  scale_fill_manual(values = c("#ffe5ab", "#cce8f8")) +
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  theme_bw()

grid.arrange(p1, p2, ncol = 2)

# summary statistics by response level
by(df[, c("covariate1", "covariate2")], df$response, summary)
```

**Comment template (continuous covariates):**

Since all covariates in this study are continuous, the descriptive analysis focuses on numerical summaries and boxplots by response level — the analogue of contingency tables for categorical covariates.

Looking at `[covariate1]`, subjects with `[success_event]` show, on average, [higher / lower / similar] values than those with `[failure_event]`. In particular, the median of `[covariate1]` is `[value_success]` for the `[success]` group and `[value_failure]` for the `[failure]` group, suggesting that [interpretation in context]. The interquartile range is also `[wider / narrower]` in the `[success / failure]` group, indicating [more / less] variability in this subgroup.

Regarding `[covariate2]`, the medians are `[similar / different]` between groups (`[value]` vs `[value]`), but the distributions differ in `[describe: shape / spread / tails]`. Specifically, the first quartile is `[Q1_success]` for `[success]` and `[Q1_failure]` for `[failure]`, suggesting that [interpretation].

These preliminary observations suggest that `[covariate1]` is likely to be [more / less] strongly associated with the response than `[covariate2]`. This hypothesis will be formally assessed via the logistic regression model.

### Case B: Binary covariates

```{r eda-binary, eval=FALSE}
# contingency tables (absolute frequencies)
table(df$response, df$binary_cov1)

# relative frequencies (by column = within each level of the covariate)
round(prop.table(table(df$response, df$binary_cov1), margin = 2), 3)

# barplot for visualization
ggplot(df, aes(x = binary_cov1, fill = response)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  xlab("[binary covariate label]") +
  scale_fill_manual(values = c("#ffe5ab", "#cce8f8")) +
  theme_bw() +
  ggtitle("[Response] by [binary covariate]")
```

**Comment template (binary covariates):**

The contingency table for `[response]` vs `[binary_cov1]` shows that `[n_success_in_group1]` out of `[total_group1]` subjects with `[binary_cov1] = 1` have `[success_event]` (`[X]%`), compared to `[n_success_in_group0]` out of `[total_group0]` (`[Y]%`) in those with `[binary_cov1] = 0`. The substantial [difference / similarity] in proportions suggests that `[binary_cov1]` is [likely / unlikely] to be associated with the response.

A preliminary inspection thus suggests that `[binary_cov1]` and `[binary_cov2]` [show / do not show] association with `[response]`, but the magnitude and statistical significance of each effect will be formally assessed via the logistic regression model, which also accounts for the joint effect of all covariates simultaneously.

### Case C: Mixed (binary + continuous)

Use this opening sentence:

Since the available covariates are of mixed types — `[list binary]` are binary indicators while `[list continuous]` are continuous variables — the exploratory analysis uses different tools for each: contingency tables for the binary covariates and boxplots/numerical summaries by response level for the continuous ones.

Then write a paragraph for each covariate using the appropriate template from Cases A or B above.

---

## Step 3 — What are the logits (template)

In this study, the logit represents the logarithm of the odds of [SUCCESS_EVENT] versus [FAILURE_EVENT], modeled as a linear function of the available covariates:

$$
\text{logit}(p_i) = \log\left(\frac{p_i}{1 - p_i}\right) 
= \log\left(\frac{P(Y_i = \text{[success_label]})}{P(Y_i = \text{[failure_label]})}\right)
$$

where $p_i = P(Y_i = \text{[success_label]})$ is the probability that individual $i$ [describe success event in plain language].

The logistic regression model assumes the logit to be a linear function of the covariates [list covariates]:

$$
\text{logit}(p_i) = \beta_0 + \beta_1 \cdot \text{[covariate1]}_i + \beta_2 \cdot \text{[covariate2]}_i + \dots
$$

While probabilities $p_i$ are bounded in $[0, 1]$ and odds $p_i / (1 - p_i)$ in $[0, +\infty)$, the logit transformation maps these quantities to the entire real line $(-\infty, +\infty)$, making linear modeling with arbitrary covariate values appropriate. The estimated coefficients $\hat\beta_k$ describe how the log-odds of [success_event] change with a one-unit increase in the corresponding covariate, holding all others constant.

---

## Step 4 — Fit del modello

```{r logit-fit, eval=FALSE}
mod1 <- glm(y ~ x1 + x2 + x3,
            data   = df,
            family = binomial)

summary(mod1)
```

---

## Step 5 — Lettura del summary

```
Coefficients:
              Estimate Std. Error z value Pr(>|z|)
(Intercept)   -1.234     0.456    -2.71   0.0067 **    ← Wald test (z, non t)
x1             0.892     0.234     3.81   0.0001 ***   ← significativo
x2            -0.456     0.298    -1.53   0.1262        ← non significativo

    Null deviance: 320.5  on 199  degrees of freedom   ← deviance modello nullo
Residual deviance: 245.8  on 196  degrees of freedom   ← deviance modello stimato
AIC: 253.8
```

**Cosa commentare:**

- **Wald z-test** su ogni coefficiente (non t-test): valore critico al 5% è $|z| > 1.96$
- **Null deviance**: deviance del modello solo-intercetta
- **Residual deviance**: deviance del modello stimato
- La riduzione da null a residual deviance misura il miglioramento dovuto alle covariate
- **AIC** per confronto tra modelli

### Comment template — summary completo

**Block 1 — Estimated model equation**

The estimated logistic regression model is:

$$
\text{logit}(\hat p_i) = [\hat\beta_0] + [\hat\beta_1] \cdot \text{[var1]}_i + [\hat\beta_2] \cdot \text{[var2]}_i + \dots
$$

**Block 2 — Intercept comment**

The **intercept** ($\hat\beta_0 = [value]$) is [statistically significant / not statistically significant] at the [α]% level (z = [value], p-value = [value]). It represents the log-odds of [success_event] for an individual with all covariates equal to zero. Since this combination is [observed / outside the observed range / not meaningful in context], the intercept [has / does not have] a substantive interpretation.

**Block 3 — Single coefficient interpretation (repeat for each covariate)**

The coefficient of **`[covariate]`** ($\hat\beta_k = [value]$) is [positive / negative] and **[statistically significant at the [α]% level / not statistically significant]** (z = [value], p-value = [value]).

This indicates that `[covariate]` is [positively / negatively] associated with the probability of [success_event]: [higher / lower] values of `[covariate]` correspond to [higher / lower] odds of [success_event], holding the other covariates constant. The corresponding odds ratio is $\exp([value]) \approx [OR]$, meaning that a one-unit increase in `[covariate]` multiplies the odds by a factor of [OR] — equivalently, the odds [increase / decrease] by approximately [|OR-1|·100]% per unit increase.

[Optional: contextualize using the natural range of the variable]

Although this effect appears [small / moderate / large] per unit, the natural range of `[covariate]` in the sample ([range]) implies a substantial cumulative effect across the observed values.

**Block 4 — For non-significant coefficients**

The coefficient of `[covariate]` ($\hat\beta_k = [value]$) is **not statistically significant** at conventional levels (z = [value], p-value > 0.05). After controlling for the other covariates in the model, `[covariate]` does not appear to contribute significantly to explaining the response. This may reflect [multicollinearity with another covariate / limited variation in the sample / a genuinely small effect that the current sample size cannot detect].

**Block 5 — Goodness of fit**

Regarding the **goodness of fit**, the null deviance is $[null\_dev]$ on $[df\_null]$ degrees of freedom, while the residual deviance is $[res\_dev]$ on $[df\_res]$ degrees of freedom — a reduction of $[null\_dev - res\_dev]$ on $[df\_null - df\_res]$ degrees of freedom. This reduction suggests that the included covariates [substantially / modestly / marginally] improve the fit compared to the null model (intercept only). The AIC of the fitted model is $[AIC]$.

**Block 6 — Convergence note**

The Fisher Scoring algorithm converged in $[n]$ iterations, confirming the absence of numerical issues during maximum likelihood estimation.

**Block 7 — Summary and transition**

Overall, [number] out of [total] covariates show a statistically significant association with [response]: [briefly summarize signs and directions]. The next step is to [compute confidence intervals for the odds ratios / perform model selection / evaluate predicted probabilities].

---

## Step 6 — Model selection con `stepAIC`

### AIC and BIC — teoria

The **Akaike Information Criterion (AIC)** is a criterion used for model selection that balances the goodness of fit of a model with its complexity. It is defined as:

$$
\text{AIC} = -2 \log L(\hat\theta) + 2k
$$

where $\log L(\hat\theta)$ is the maximized log-likelihood of the model and $k$ is the number of estimated parameters. The first term rewards models that fit the data well (higher log-likelihood means lower $-2\log L$), while the second term penalizes models with many parameters, discouraging overfitting.

Lower values of AIC indicate a better trade-off between fit and parsimony. AIC is **always used to compare models on the same dataset**, never interpreted in absolute terms. Among a set of candidate models, the one with the smallest AIC is preferred.

AIC is derived from information theory: it is an estimator of the expected Kullback-Leibler divergence between the fitted model and the true (unknown) data-generating process. Its objective is therefore predictive — it aims to identify the model that would best predict new, unseen data.

A closely related criterion is the **Bayesian Information Criterion (BIC)**, defined as $-2 \log L(\hat\theta) + k \log n$. The penalty is larger than that of AIC whenever $\log n > 2$ (i.e. for any sample size $n > 7$, virtually always in practice), so BIC tends to select more parsimonious models. The two criteria have different philosophical foundations: BIC asymptotically identifies the "true" model (if one exists in the candidate set), while AIC is asymptotically optimal for prediction.

### Codice R

```{r logit-stepwise, eval=FALSE}
library(MASS)

# selezione stepwise basata su AIC (default)
mod_step <- stepAIC(mod_full, direction = "both", trace = TRUE)

# con BIC invece di AIC
n <- nrow(df)
mod_step_bic <- stepAIC(mod_full, direction = "both", k = log(n))

summary(mod_step)
```

### Lettura del trace output

```
Start:  AIC=99.38
r ~ osmo + cond

         Df Deviance    AIC
<none>         93.38  99.38     ← current model
- cond    1   99.41 103.41      ← removing cond would raise AIC
- osmo    1  103.05 107.05      ← removing osmo would raise AIC even more
```

- La riga `<none>` mostra il modello corrente e il suo AIC
- Ogni riga `- variable` mostra cosa succederebbe rimuovendo quella variabile
- Se qualche riga `- variable` ha AIC più basso di `<none>`, quella variabile viene rimossa
- La procedura si ferma quando nessuna rimozione o aggiunta migliora l'AIC

### Comment template — modello finale = modello pieno

Stepwise selection based on AIC was performed starting from the full model containing `[list initial covariates]`. The procedure considered both forward and backward steps at each iteration. The final selected model includes all initial covariates `[list]` with AIC = $[value]$, indicating that no removal of any covariate would improve the criterion. This suggests that all covariates contribute meaningfully to the model fit, in agreement with the significance levels of the individual Wald tests reported in the previous summary.

### Comment template — variabili rimosse

Stepwise selection based on AIC was performed starting from the full model with `[p]` covariates: `[list initial covariates]`. At each iteration the procedure evaluated both forward and backward moves, retaining the candidate that minimized the AIC.

At step 1, the variable `[removed_var]` was removed, lowering the AIC from $[AIC_before]$ to $[AIC_after]$. At step 2, `[next_removed]` was further removed, [continue as needed]. The procedure terminated when no further removal or addition improved the criterion.

The **final selected model** includes `[final covariates]` with AIC = $[AIC_final]$, considerably lower than the initial AIC of $[AIC_full]$. The removed covariates — `[removed list]` — did not provide enough explanatory power to justify their inclusion in the model.

This result is consistent with the Wald tests reported in the full model summary: the removed covariates had p-values [larger than conventional thresholds / inconsistent significance], suggesting limited individual contributions after controlling for the others.

### Summary del modello finale

After selection, refit / inspect the final model:

```{r final-model, eval=FALSE}
summary(mod_step)
```

The summary of the selected model is reported above. The coefficients are [describe: all significant / a mix of significant and not]. The AIC ($[final\_AIC]$) confirms the gain in parsimony obtained from the selection. The estimated equation of the final model is:

$$\text{logit}(\hat p_i) = [\hat\beta_0] + [\hat\beta_1] \cdot \text{[var1]} + \dots$$

The next steps will be to interpret each coefficient via odds ratios with confidence intervals, and to compute predicted probabilities for new observations.

---

## Step 7 — Confidence intervals e odds ratio

In regressione logistica, "confidence interval" può riferirsi a **due quantità diverse** a seconda della scala. Specifica sempre quale stai usando.

### CI sulla scala logit (per $\beta_k$)

L'intervallo di valori plausibili per il **vero coefficiente sulla scala dei log-odds**.

$$
\hat\beta_k \pm 1.96 \cdot SE(\hat\beta_k)
$$

```{r ci-logit, eval=FALSE}
confint.default(mod)     # asymptotic Wald CI (fast)
confint(mod)              # profile-likelihood CI (more accurate, slower)
```

Raramente commentato in isolamento — i log-odds sono poco interpretabili.

### CI sulla scala dell'odds ratio (per $\exp(\beta_k)$) — quello da commentare

L'intervallo di valori plausibili per l'**effetto moltiplicativo sugli odds**.

$$
\left[\exp(\hat\beta_k - 1.96 \cdot SE),\; \exp(\hat\beta_k + 1.96 \cdot SE)\right]
$$

```{r ci-or, eval=FALSE}
exp(confint.default(mod))      # OR CI (Wald)
exp(confint(mod))               # OR CI (profile likelihood)

# clean table with point estimate + CI
cbind(OR = exp(coef(mod)), exp(confint.default(mod)))
```

### Interpretation table

| Observation | Conclusion |
|-------------|------------|
| CI for OR excludes 1 | Effect is statistically significant at 5% |
| CI for OR includes 1 | Effect is NOT significant (equivalent to p > 0.05) |
| CI for OR very wide | Imprecise estimate (small sample / multicollinearity) |
| CI for OR includes very different values (e.g. [0.5, 5]) | Direction of effect uncertain |

### Perché 1.96 (non il valore critico della t)

Unlike linear regression, where the test statistic follows Student's t with $n-p-1$ degrees of freedom, the Wald test in logistic regression follows the standard normal distribution **asymptotically**. The 95% critical value is therefore exactly 1.96 (not approximately). For small $n$ or sparse data, this asymptotic approximation can be unreliable; in those cases, prefer profile-likelihood CIs via `confint(mod)` or use likelihood ratio tests via `anova(mod0, mod1, test = "Chisq")`.

### Regole di interpretazione degli OR

| Quantità | Cosa rappresenta |
|----------|--------------------|
| $\exp(\hat\beta_0)$ | Odds di successo quando tutte le covariate sono zero |
| $\exp(\hat\beta_k)$ per $x_k$ continua | Cambiamento **moltiplicativo** degli odds per un aumento di un'unità in $x_k$, a parità di covariate |
| $\exp(\hat\beta_k)$ per $x_k$ binaria | Rapporto tra gli odds dei due gruppi, a parità di covariate |
| $\exp(\hat\beta_k) > 1$ | $x_k$ **aumenta** gli odds di successo |
| $\exp(\hat\beta_k) < 1$ | $x_k$ **riduce** gli odds di successo |
| $\exp(\hat\beta_k) = 1$ | Nessun effetto (CI che include 1 → non significativo) |

### Conversione rapida: variazione percentuale degli odds

$$\text{variazione \%} = (\exp(\hat\beta_k) - 1) \cdot 100\%$$

- $\exp(\hat\beta_k) = 1.05$ → "gli odds aumentano del 5%"
- $\exp(\hat\beta_k) = 1.50$ → "gli odds aumentano del 50%"
- $\exp(\hat\beta_k) = 0.80$ → "gli odds diminuiscono del 20%"

### Comment template — single odds ratio

The 95% confidence interval for the odds ratio of `[variable]` is `[L, U]`. With 95% confidence, the true odds ratio lies in this range. Since the interval [does not include / includes] 1, the effect of `[variable]` on the odds of [success_event] is [statistically significant / not significant] at the 5% level. Specifically, a one-unit increase in `[variable]` multiplies the odds of [success] by a factor between `[L]` and `[U]`, holding other covariates constant — equivalent to a [decrease/increase] of `[(1-L)·100]%` to `[(U-1)·100]%`.

### Comment template — full coefficient discussion

The variable `[name]` has an estimated coefficient $\hat\beta = [value]$ on the logit scale, with z-statistic [value] and p-value [value]. The corresponding odds ratio is $\exp(\hat\beta) = [OR]$ with 95% CI $[L, U]$.

Interpretation: [a one-unit increase in / membership in the reference group of] `[name]` is associated with a multiplicative change in the odds of [success] by a factor of [OR], holding other covariates constant. The CI [does not / does] include 1, confirming [the statistical significance / the lack of significance] of this effect at the 5% level.

### Comment template — binary covariate

The estimated odds ratio for `[group1]` vs `[group0]` is $\exp(\hat\beta) = [value]$. Holding other covariates constant, subjects in `[group1]` have odds of [outcome] that are [value] times those in `[group0]`. The 95% CI is $[L, U]$, [not] including 1, confirming [the presence / the absence] of a significant association.

---

## Step 8 — Predizione delle probabilità

```{r logit-predict, eval=FALSE}
# nuova osservazione
new_data <- data.frame(x1 = 25, x2 = "level_B", x3 = 1)

# probabilità stimata di successo (SEMPRE type = "response")
predict(mod_final, newdata = new_data, type = "response")

# probabilità in-sample per tutto il dataset
fitted_probs <- predict(mod_final, type = "response")

# classificazione con soglia 0.5 e matrice di confusione
predicted_class <- ifelse(fitted_probs > 0.5, 1, 0)
table(predicted = predicted_class, observed = df$y)

# per prendere prime x
fitted_probs[1:x]
```

| Tipo | Output | Comando R |
|------|--------|-----------|
| Logit | $\hat{\text{logit}}(p_i) = x'_i \hat\beta$ | `predict(mod, newdata)` (default) |
| Probabilità | $\hat{p}_i \in [0, 1]$ | `predict(mod, newdata, type = "response")` |

**Comment template:**

For a new unit with characteristics `[describe values]`, the estimated probability of `[success_event]` is $\hat{p} = [value]$. Under a 0.5 classification threshold, the unit is predicted as `[class 1 / class 0]`, indicating that the model considers it `[likely / unlikely]` to be a success.

---

## Step 9 (opzionale) — Likelihood Ratio Test per modelli annidati

```{r lrt, eval=FALSE}
mod_reduced <- glm(y ~ x1, data = df, family = binomial)
mod_full    <- glm(y ~ x1 + x2 + x3, data = df, family = binomial)

# LRT (chi-quadro): usare "Chisq", NON F-test come in lm()
anova(mod_reduced, mod_full, test = "Chisq")
```

La statistica è la **differenza di deviance** $D = 2[\ell(M_1) - \ell(M_0)]$, distribuita come $\chi^2_{p_1 - p_0}$ sotto $H_0$.

**Comment template:**

The likelihood ratio test comparing the reduced model (`[M0 covariates]`) to the full model (`[M1 covariates]`) yields a deviance difference of [value] on [df_diff] degrees of freedom (p-value [value]). We [reject / fail to reject] $H_0$ that the additional covariates have no effect: at the [α]% level, `[var_x]` and `[var_y]` [jointly contribute / do not jointly contribute] to explaining the response, beyond what is already captured by the reduced model.

---

## Trappole comuni (regressione logistica)

- **Wald test, non t-test**: il test su ogni coefficiente in `summary(glm)` è Wald con riferimento $\mathcal{N}(0,1)$. Valore critico al 5%: 1.96 esatto
- **I coefficienti sulla scala logit non si interpretano**: esponenziare sempre per ottenere odds ratio
- **CI per OR**: prendere `exp()` dell'CI del coefficiente. Non calcolare SE dell'OR direttamente
- **Livello di riferimento**: R sceglie il primo livello alfabetico. Usare `relevel()` per controllare
- **Per confronto modelli annidati usare $\chi^2$, non F**: `anova(M0, M1, test = "Chisq")`
- **CI che include 1**: se l'CI per un OR include 1, la variabile NON è significativa
- **Predire probabilità**: usare sempre `type = "response"`, altrimenti si ottengono logit

## Confronto rapido: `lm` vs `glm`

| Scenario | `lm` (lineare) | `glm` (logistico) |
|----------|----------------|-------------------|
| Test singolo coefficiente | t-test (esatto) | Wald test (asintotico, normale) |
| Valore critico 5% | $t^*_{n-p-1, 0.975}$ (≈ 1.96 per $n$ grande) | 1.96 esatto |
| CI singolo coefficiente | $\hat\beta \pm t^* \cdot SE$ | $\hat\beta \pm 1.96 \cdot SE$ |
| Test globale | F-test (in summary) | Confronto null vs residual deviance |
| Confronto modelli annidati | `anova(M0, M1)` → F-test | `anova(M0, M1, test = "Chisq")` → LRT |
| Selezione automatica | `stepAIC(...)` | `stepAIC(...)` (identico) |
| Predizione | `predict(mod, newdata)` | `predict(mod, newdata, type = "response")` |


# Clustering Model-Based (Ch. 13)

## Quando usarlo

Usa il clustering model-based quando:

- Hai **dati continui senza etichette** (setting non supervisionato)
- Vuoi **identificare sottogruppi** (cluster) con struttura distributiva potenzialmente diversa
- Vuoi un approccio **probabilistico**: ogni unità ha una probabilità posteriore di appartenere a ogni cluster (non solo assegnazione hard)

I dati si assumono provenienti da una **mistura finita di distribuzioni gaussiane**:

$$f(x) = \sum_{k=1}^{K} \pi_k \cdot \mathcal{N}_p(x; \mu_k, \Sigma_k)$$

dove $K$ è il numero di cluster, $\pi_k$ sono i pesi della mistura ($\sum_k \pi_k = 1$), e ogni cluster ha la propria media $\mu_k$ e covarianza $\Sigma_k$.

---

## Il framework mclust

Il pacchetto `mclust` parametrizza le matrici di covarianza via decomposizione spettrale:

$$\Sigma_k = \lambda_k D_k A_k D_k^\top$$

- $\lambda_k$: scalare — controlla il **volume** del cluster
- $D_k$: matrice ortogonale — controlla l'**orientazione**
- $A_k$: matrice diagonale — controlla la **forma**

I modelli sono codificati con **3 lettere** (volume, forma, orientazione):

| Lettera | Significato |
|---------|---------|
| E | Uguale tra cluster |
| V | Variabile tra cluster |
| I | Identità / sferico |

| Codice | Volume | Forma | Orientazione | Descrizione geometrica |
|--------|--------|-------|-------------|------------------------|
| **EII** | E | I | I | Cluster sferici, stessa dimensione |
| **VII** | V | I | I | Cluster sferici, dimensioni diverse |
| **EEI** | E | E | I | Cov. diagonale, allineata agli assi, stessa forma e volume |
| **VVI** | V | V | I | Cov. diagonale, allineata agli assi, forme e volumi diversi |
| **EEE** | E | E | E | Stesso ellissoide per tutti (solo le medie differiscono) |
| **VVV** | V | V | V | Completamente flessibile — ogni cluster ha il proprio ellissoide |

> **Trade-off**: da EII (più parsimonioso) a VVV (più flessibile). VVV può overfittare con $K$ grande o $n$ piccolo.

---

## Fit del modello con mclust

```{r mclust-fit, eval=FALSE}
library(mclust)

# selezione automatica di K e modello basata su BIC
mod_clust <- Mclust(df, G = 1:9,, modelNames = "e/v")

# output principali
summary(mod_clust)              # info sintetica: modello, K, log-lik, BIC
mod_clust$G                     # numero di cluster selezionato
mod_clust$modelName             # codice modello (es. "VVV", "EEE")
mod_clust$BIC                   # valori BIC per tutte le combinazioni K × modello
mod_clust$classification        # assegnazione cluster per ogni unità (regola MAP)
mod_clust$z                     # probabilità posteriori di appartenenza (matrice n × K)
mod_clust$parameters$mean       # medie stimate per cluster
mod_clust$parameters$variance   # struttura di covarianza stimata
```

### Lettura dell'output di summary

```
Mclust VVV (ellipsoidal, varying volume, shape, and orientation) model with 3 components:

 log-likelihood   n df       BIC       ICL
      -1234.56 200 27   -2598.6   -2615.3

Clustering table:
 1   2   3
65  92  43
```

- **Modello selezionato**: VVV con 3 cluster
- **log-likelihood**: log-verosimiglianza massima alla convergenza dell'EM
- **df**: numero di parametri stimati
- **BIC**: criterio di selezione (in mclust **più alto è meglio** — convenzione opposta al solito)
- **ICL**: Integrated Completed Likelihood (penalizza l'overlap tra cluster)
- **Clustering table**: quante unità in ogni cluster

### Comment template — mclust summary

The summary of the model reports the value of the **log-likelihood at 
convergence** (equal to `[log-lik value]`), the **BIC** and **ICL** indices 
(`[BIC value]` and `[ICL value]`), the **number of observations** (`[n]`) 
and the **degrees of freedom** — computed as the number of free parameters 
estimated by the model (`[df]`).

The output also provides the **number of observations classified in each 
cluster** by the MAP rule. We observe that `[cluster A]` is `[larger / 
smaller / similarly sized]` than `[cluster B]`, with `[n_A]` subjects 
compared to `[n_B]`.

[Optional, one closing sentence if relevant:]

The selected model is **`[model code, e.g., VVV / EEE]`** with **`[K]` 
components**, indicating that the data are best described by clusters with 
`[describe geometric structure: e.g., "varying volume, shape and orientation" 
/ "equal volume, shape and orientation"]`.

### Parameters = TRUE 
The parameters of the estimated mixture model are:

**(i)** the **mixing probabilities (weights)** of the `[K]` mixture components,
**(ii)** the **means** of the components,  
**(iii)** the **variances** (or covariance matrices, in the multivariate case).

**Mixing weights**: The `[K]` components have weights of `[π_1, π_2, ...]`, 
respectively. Note that these values represent the **estimated proportions 
of the population** belonging to each component — they do not necessarily 
coincide with the empirical proportions of subjects allocated to each group 
by the MAP rule.

**Means**: The mean value of `[variable name]` for subjects in component 1 
is `[μ_1]`, `[much lower / similar / higher]` than the corresponding value 
for component 2 (`[μ_2]`). We can therefore characterize the `[K]` 
subpopulations based on the value of `[variable]`: `[describe direction: 
higher / lower]` values on average for `[component A]`, `[higher / lower]` 
for `[component B]`.

**Variances**: The variance values for the components are `[equal / 
different]`, consistent with the model code `[model_code, e.g., EEE, VVV]` 
selected. The corresponding standard deviation is approximately `[σ]`; 
this value represents the average variability around the mean of each 
component.


---

## Visualizzazione

```{r mclust-plot, eval=FALSE}
# plot multiplo: BIC, classificazione, incertezza, densità
plot(mod_clust, what = "BIC")              # BIC vs K per ogni modello
plot(mod_clust, what = "classification")   # scatterplot colorato per cluster
plot(mod_clust, what = "uncertainty")      # incertezza nelle assegnazioni
plot(mod_clust, what = "density")          # contorni della densità della mistura
```

### Interpretazione del BIC plot

The BIC plot shows BIC values for all combinations of $K$ (x-axis) and model code (different lines). The optimal combination has the **highest BIC** (note: mclust reports BIC with the opposite sign convention — higher is better).

**Comment template:**

> The BIC plot indicates that the model `[code]` with $K = [value]$ clusters maximizes the BIC. Alternative models with similar BIC values include `[alternatives]`, suggesting that the cluster structure is [robust to / sensitive to] the choice of parameterization.

---

## Interpretazione e validazione dei cluster

```{r cluster-stats, eval=FALSE}
# dimensioni dei cluster
table(mod_clust$classification)

# medie per cluster
mod_clust$parameters$mean

# covarianze per cluster (per VVV: Sigma separata per ogni cluster)
mod_clust$parameters$variance$sigma

# descrivere ogni cluster su ogni variabile
df$cluster <- mod_clust$classification
aggregate(df[, -ncol(df)], by = list(df$cluster), FUN = mean)

# incertezza nelle assegnazioni
unc <- 1 - apply(mod_clust$z, 1, max)
high_unc <- which(unc > 0.3)   # unità al confine tra cluster
length(high_unc)
```

**Comment template:**

> The model selected by mclust is `[code]` with $K = [value]$ clusters. Cluster sizes are: cluster 1 — [n1] units, cluster 2 — [n2] units, [...]. The clusters differ primarily in `[variable A]` (cluster 1 mean: [v1], cluster 2 mean: [v2]) and in `[variable B]` (cluster 1 mean: [v1], cluster 2 mean: [v2]).
>
> The geometric structure `[code description, e.g., VVV = ellipsoids with varying volume, shape, and orientation]` indicates that the clusters [have similar / have different] internal structure. [Optional: [X]% of units show high uncertainty in their assignment (> 0.3), suggesting they lie on cluster boundaries.]

---

## Confronto BIC vs ICL

| Criterio | Cosa premia | Quando usarlo |
|----------|------------|---------------|
| **BIC** | Verosimiglianza + parsimonia | Scopo predittivo / descrittivo |
| **ICL** | Come BIC, ma penalizza l'overlap tra cluster | Quando si vuole separazione netta (classificazione) |

Se due modelli hanno BIC simile ma ICL molto diverso, significa che uno ha cluster molto più sovrapposti dell'altro.

---

## Trappole comuni (clustering)

- **Standardizzare le variabili prima** del clustering se hanno scale molto diverse — altrimenti la variabile con il range maggiore domina la struttura di $\Sigma$

  ```{r scale-before-mclust, eval=FALSE}
  df_std <- scale(df)
  mod <- Mclust(df_std)
  ```
- **Convenzione BIC di mclust**: in mclust **BIC più alto è meglio** (opposto rispetto a `lm`/`glm`). Non confondersi
- **L'EM può convergere a massimi locali**: mclust testa automaticamente più inizializzazioni, ma è bene saperlo
- **Non pre-specificare K rigidamente**: usare `Mclust(df, G = 1:9)` per esplorare
- **VVV può overfittare con pochi dati**: con poche osservazioni per cluster, preferire EEE o EEI per evitare stime di varianza instabili
- **Variabili categoriali**: `mclust` è progettato per variabili continue. Per dati misti usare solo le continue, o valutare approcci alternativi

---

### Step 5 — Classification of each individual (Es. 32 point .5, Es. 33 point .4)

```{r mclust-classification, eval=FALSE}
# classification by MAP rule
mod_clust$classification
table(mod_clust$classification)
prop.table(table(mod_clust$classification))

# posterior probabilities of cluster membership (n x K)
head(mod_clust$z)

round(head(cbind(mod_clust$z, Cluster = mod_clust$classification)), 4)

```

**Comment template — classification:**

The classification of each individual is obtained by applying the **MAP rule 
(Maximum A Posteriori)**: each unit is assigned to the cluster with the 
highest estimated posterior probability of membership. Specifically:

$$
\hat z_i = \arg\max_{k = 1, \dots, K} \hat p_{ik}
$$

where $\hat p_{ik}$ is the estimated posterior probability that unit $i$ 
belongs to cluster $k$, computed via Bayes' rule using the estimated 
mixing weights and component densities.

The resulting classification produces `[K]` groups with sizes `[n_1, n_2, ...]` 
(empirical proportions: `[p_1, p_2, ...]`). These empirical proportions are 
`[close to / different from]` the estimated mixing weights `[π_1, π_2, ...]`, 
as expected: the mixing weights estimate the **population-level** 
proportions, while the empirical proportions reflect the **sample-level** 
classification.

**Applicative interpretation**: 

For the application at hand, this classification allows us to `[describe: 
"distinguish between two latent subpopulations of patients based on their 
haemoglobin levels, potentially corresponding to clinical categories such 
as anaemic vs healthy individuals" / "group fishes into age classes based 
on length, useful for fishery management without requiring direct age 
measurements"]`. The model thus provides a data-driven, unsupervised 
classification of subjects without needing pre-existing labels.

---

### Step 6 — Visualizing the clustering partition (Es. 32 point .6, Es. 33 point .5)

#### Univariate case (most common in Es. 32-33)

```{r mclust-class-plot-univariate, eval=FALSE}
# main multi-plot
plot(mod_clust, what = "classification")

# customized stripchart-style univariate visualization
stripchart(haemoglobin ~ mod_clust$classification,
           method = "jitter", pch = 16,
           col = c("salmon", "steelblue"),
           main = "Cluster partition",
           xlab = "Haemoglobin")
legend("topright",
       legend = c("Cluster 1", "Cluster 2"),
       col = c("salmon", "steelblue"),
       pch = 16, cex = 0.8)
```

#### Multivariate case (occasional, for completeness)

```{r mclust-class-plot-multivariate, eval=FALSE}
# pairwise scatterplots colored by cluster
plot(mod_clust, what = "classification")
```

**Comment template — classification plot:**

The plot shows the partition of the sample into `[K]` clusters, with each 
unit colored according to its MAP assignment. The clusters are 
`[clearly separated / partially overlapping / heavily overlapping]` along 
the `[variable]` axis: cluster 1 is concentrated in the range `[range_1]`, 
while cluster 2 occupies the range `[range_2]`. The boundary between 
clusters lies around `[boundary value]`, which corresponds to the value at 
which the posterior probabilities of the two components are approximately 
equal.

[Optional, for Es. 33 with K=4]:

For the snapper dataset with 4 components, the clusters partition the length 
distribution into four groups, plausibly corresponding to four different age 
classes. The mean lengths increase monotonically across clusters, consistent 
with progressive fish growth: cluster 1 (mean = `[μ_1]`) likely represents 
the youngest fish, while cluster 4 (mean = `[μ_4]`) represents the oldest. 
The clusters with intermediate means (`[clusters 2 and 3]`) show 
`[some / substantial]` overlap, reflecting the natural continuity of growth.

---

### Step 7 — Mixture density + histogram (Es. 32 point .7, Es. 33 point .6)

This is one of the most distinctive plots of mixture analysis: shows visually 
how well the mixture density fits the observed data.

```{r mclust-density-plot, eval=FALSE}
# grid of values over the range of the data
x_grid <- seq(min(haemoglobin), max(haemoglobin), length.out = 500)

# extract estimated parameters
pi_hat    <- mod_clust$parameters$pro
mu_hat    <- mod_clust$parameters$mean
sigma_hat <- sqrt(mod_clust$parameters$variance$sigmasq)  # for univariate

# evaluate the mixture density on the grid
mixture_density <- numeric(length(x_grid))
for (k in 1:length(pi_hat)) {
  mixture_density <- mixture_density + 
                     pi_hat[k] * dnorm(x_grid, mean = mu_hat[k], sd = sigma_hat[k])
}

# plot histogram with overlaid mixture density
hist(haemoglobin,
     freq = FALSE,                       # density scale, not frequencies
     breaks = 20,
     col = "lightgray", border = "white",
     main = "Histogram and estimated mixture density",
     xlab = "Haemoglobin", ylab = "Density")
lines(x_grid, mixture_density,
      col = "red", lwd = 2)

# also plot individual component densities
for (k in 1:length(pi_hat)) {
  lines(x_grid, 
        pi_hat[k] * dnorm(x_grid, mean = mu_hat[k], sd = sigma_hat[k]),
        col = "blue", lty = 2, lwd = 1.5)
}
legend("topright",
       legend = c("Mixture density", "Component densities"),
       col = c("red", "blue"),
       lty = c(1, 2), lwd = 2, cex = 0.7)
```

**Comment template — mixture density + histogram:**

The figure shows the histogram of the observed values of `[variable]` 
together with the estimated mixture density (solid line) and the individual 
component densities (dashed lines), weighted by their respective mixing 
probabilities.

The mixture density `[fits the histogram closely / shows a good agreement 
with the observed distribution / shows some discrepancy in certain regions]`. 
In particular, the `[bimodal / multimodal]` shape of the histogram is 
captured by the mixture: each peak in the histogram corresponds to one of 
the `[K]` component densities, suggesting that the observed heterogeneity 
is well explained by `[K]` distinct subpopulations.

The individual component densities reveal the structure underlying the 
mixture: component 1 (centered at `[μ_1]`) captures the `[describe: e.g., 
"left peak of the distribution"]`, while component 2 (centered at `[μ_2]`) 
captures the `[describe]`. The valleys between peaks correspond to regions 
of low density in the data, where membership in one of the two clusters is 
uncertain.

This visual confirmation supports the choice of `[K]` components: a 
unimodal model (e.g., single Gaussian) would fail to capture the observed 
multimodality, while a model with more components would likely overfit.

---

### Step 8 — Prediction for new units (Es. 33 point .7 only)

```{r mclust-predict, eval=FALSE}
# new fish with length = 7
new_data <- data.frame(snapper = 7)

# predict using the predict.Mclust method
pred <- predict(mod_clust, newdata = new_data)

# returns:
# - classification: assigned cluster
# - z: posterior probabilities of membership in each cluster
pred$classification
pred$z
```

**Comment template — prediction for a new unit:**

For a new fish with length equal to `[value]` inches, the model assigns it 
to cluster `[N]` based on the MAP rule, with the highest estimated posterior 
probability `[max p]`. The complete vector of posterior probabilities is:

| Cluster | Posterior probability |
|---------|----------------------|
| 1 | `[p_1]` |
| 2 | `[p_2]` |
| 3 | `[p_3]` |
| 4 | `[p_4]` |

The uncertainty of this prediction is $1 - \max_k \hat p_k = `[uncertainty]`$. 
A `[low / high]` uncertainty indicates that the new unit is `[clearly / 
ambiguously]` assigned to its cluster, lying `[centrally within / near the 
boundary of]` the cluster's region.

(uncertanty: <0.05 very low - 0.10 low - 0-20 medium > 0.3 high)

**Applicative interpretation**: 

A snapper of length 7 inches is most likely a `[young / adult / intermediate]` 
fish, consistent with the age class represented by cluster `[N]`. This kind 
of prediction is useful for fishery management: given only length 
measurements (which are easy and inexpensive to collect), the model 
provides a probabilistic estimate of the age class, which would otherwise 
require destructive sampling (e.g., otolith analysis).

---

# Flowchart Decisionale

<!-- Giorno 15: guida "dato un problema X, quali passi?" -->

## Schema rapido

- **Risposta continua** → `lm()`
  - Verifica assunzioni: `plot(model)` (4 diagnostic plots)
  - Controlla multicollinearità: `vif(model)`
  - Troppe covariate: `stepAIC(model)`
  - Predizione: `predict(..., interval = "confidence" | "prediction")`
- **Risposta binaria** → `glm(family = binomial)`
  - Interpretazione: `exp(coef(model))` per odds ratio
  - Predizione probabilità: `predict(..., type = "response")`
- **CI per statistica non standard** → bootstrap (manuale o `boot::boot`)
- **Correlazione "sparisce" condizionando su Z** → correlazione parziale (`ppcor::pcor`)

---

# Trappole Comuni

<!-- Annotare qui errori frequenti durante lo studio -->

- R² aumenta sempre aggiungendo covariate (anche inutili) → usare R² adj per confronti
- F-test significativo ma nessun t-test significativo → sintomo di multicollinearità
- Leverage alto ≠ punto influente (può esserlo o no)
- Nei GLM: Wald test nel summary, likelihood ratio test via `anova(mod, test = "Chisq")`
