---
title: "LogisticRegression"
author: "me"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##29.1 descriptive statistics
```{r}

```

```{r}
table(birds$cancer, birds$female)
```
The odds of having cancer of a male

```{r}
37/74
```
The odds of having cancer of a female

```{r}
12/24
```

Males and females have the same odds of having cancer (not probability)

The odds of having cancer are the probability of having cancer divided by the probability of not having cancer.

odds ratio = odds in group 1 / odds in group2 

```{r}
0.5/0.5
```

##29.3  Fit a logistic regression model with cancer as the response variable and perform variable selection. Comment on the results obtained according to the Akaike Information Criterion. State a conclusion in context

```{r}
full_model = glm(formula = cancer ~ . , data = birds, family = binomial)
step(full_model)
```

```{r}
mod.sel = glm(formula = cancer~ yrsmoke + bird, data = birds, family = binomial)
coef(mod.sel)
```
keeping fixed the other variables, having a increase on yrsmoke gives a multiplication on the odds of having cancer of the coefficient (~6% for each year)

```{r}

```

Provide a confidence interval for the odds ratio associated with the estimated
coefficient of bird. Comment on the resul
```{r}
exp(coef(mod.sel))
```
```{r}
exp(cbind(
  OR = coef(mod.sel),
  confint(mod.sel)
))
```
The estimated odds ratio associated with bird1 is 4.37, with a 95% confidence interval equal to (2.05, 9.75).
This means that the odds of having cancer for individuals with bird = 1 are estimated to be about 4.4 times the odds for the reference group (bird = 0), holding yrsmoke constant.

Since the confidence interval does not include 1, the effect of bird is statistically significant at the 5% level.



