---
title: "Es23"
author: "me"
date: "2026-05-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, eval=FALSE}
# Caricamento librerie usate nel corso
library(MASS)       # stepAIC, Boston dataset
library(car)        # vif, diagnostica
library(corrplot)   # visualizzazione matrici di correlazione
library(boot)       # bootstrap
library(ppcor)      # correlazioni parziali
library(GGally) 
library(mvtnorm)
library(ggplot2)
library(skimr)
library(gridExtra)

```

The dataset MASS::whiteside (available in the MASS library) contains observations
on natural gas consumption (in cubic feet) and temperature (in degrees Celsius), along with an indicator of whether the house had been insulated for energy efficiency.
The data were collected over two consecutive winters, with measurements taken both before and after the insulation intervention. The 56 weekly observations are divided
into 26 weeks before and 30 weeks after the intervention.

```{r}
whiteside <- MASS::whiteside
head(whiteside)
```

### 23.1 State the research questions that can be formulated based on the available data.

The main objective is to model how the gas consumption depends on the temperature and the insulation of an house.

Specifically, we aim to:
1. Quantify the effect of temperature and insulation on gas consumption.
2. Identify which covariates are significant predictors
3. Assert wheter the insulation intervention has been an improving factor on the consumption, holding all the other covariates equal.

Based on common sense, we expect that the coefficient of the insultion variable will be negative, due to the fact that the insulation procedure usually brings benefits in the quantity of energy needed to warm.
The analysis will verify these hypotheses and explore the relationships.

### 23.2 Print the levels and labels of the categorical covariate.

```{r}
levels(whiteside$Insul)
```


```{r}
labels(whiteside$Insul)
```
The variable Insul has two different levels (Before/after) which represent wheter the misuration is done before or after the insulation procedure and the data is made up by 56 observations.

### 23.3 Provide descriptive statistics and a figure to describe the sample data. Do the data come from an observational study or form a randomized experiment? Comment on each output

The available data come from an **observational study** (not a randomized 
experiment), since the insulation intervention was applied to the same house 
at a specific point in time, without random assignment to experimental groups. 
This limits the strength of any causal interpretation: while we can quantify 
the association between insulation and consumption, we cannot exclude that 
other unmeasured factors (e.g., changes in heating behavior, different winters' 
weather patterns) contribute to the observed differences.

```{r}
skimr::skim_without_charts(whiteside)
```
```{r}
IQR(whiteside$temp)
IQR(whiteside$Gas)
```

Looking at the descriptive statistics, we see that there are 56 obseravations splitted in 30 misurations before the insulation process and 26 after. There are no missing values for this variables in the observations recorded.
The temp variable varies from -0.8 to 10.2 degrees, with an IQR of ~ 4.12 and the mean really close to the median (4.88 / 4.90), meaning that the distribution is quite symmetric around the center. 
Regarding the gas value, we can see that the values vary from 1.3 to 7.2, with an IQR of 1.125 and a SD of 1.167 meaning that the values recorded does not have a big variation (that makes sense: the observations are all coming from the same house, outliers would be strange). The fact that the mean is slightly larger than the median (4.07 vs 3.95)suggests a modest asymmetry toward higher values, although the difference is not substantial enough to indicate strong skewness.

### 23.4 Fit a multiple linear regression model to explain gas consumption as a function of the other covariates. First, use the Akaike Information Criterion to select the model. Comment on the results provided by the step function.

```{r}
mod <- lm(whiteside$Gas~ whiteside$Temp + whiteside$Insul)
mod.C = step(mod, trace = TRUE)

```
Stepwise selection based on AIC was performed starting from the full model containing both Temp and Gas covariates. The procedure considered backwards steps at each iteration. The final selected model includes all of the initial covariates with AIC = -112.316
, indicating that no removal of any covariate would improve the criterion. This suggests that all covariates contribute meaningfully to the model fit.

### 23.5 Report and comment on all the results provided by the summary function for the model selected in the previous step.

```{r}
summary(mod.C)
```
**Preliminary residual analysis**

The residuals range from $-0.74236$ to $0.74314 $ with median $0.04338$. The IQR is 
$[-0.22291, 0.24377]$. This resebles a symmetrical distribution of the residuals around the center (0), with almost symmetrical values of min/max and 1st/3rd quartiles

**The estimated model is:**

\[
\hat{Y} = 6.551 - 0.337\,\text{temp} - 1.57\,\text{insul}
\]

The intercept ($[β₀]$) does have a meaningful interpretation, since the value 0 of the covariates is realistic. As expected from our research question, the slope coefficient of the insul is negative, consistent with the hypothesis that the isolation process done to an house reduces the consumption of gas, considering fixed the value of the other covariates.

**Coefficient interpretation:**

- **Temp** ($\hat{\beta}_1 = -0.337$): significant at the 
  0% level, p-value <2e-16. A unit increase is associated with an average decrease of $0.337$ units in $Gas$, ceteris paribus.

- **Insul** ($\hat{\beta}_2 = -1.56$): significant at the 
  0% level, p-value <2e-16. It means that on average, the isulation reduces the $Gas$ consumption of $1.56$ units, holding fixed all of the other variables.

**Goodness of fit and global test**

The model explains $[R²×0.91]\%$ of the variance in $Gas$ 
($R^2 = [0.91]$, adjusted $R^2 = [0.91]$), that shows an optimal fit of the model to the data. We can also see that the R^2 and the adjusted R^2 have quite the same value and this is due to the fact that there are just 2 covariates that are both significant: for this reason, the adjusted coefficient does not penalize much more than the classical one. 
The residual standard error is $0.3574$, it means that, on average, the observed and the estimated values for each sample unit differ by 0.36. Although the value seems small, it needs to be further evaluated with respect to a centrality index (e.g., the mean) of the response variable. Computing the ratio between the RSE and the mean of the carats, we obtain the percentage error, equal to [RSE/MEAN] = 0.89 on average the percentage error between observed and fitted values is equal to 8.9%

The global F-test yields $F = 267.1$ on $p = 2$ and $53$ degrees of freedom 
(p-value $< 2.2e-16$). We reject $H_0$: at least one covariate is associated with $gas$ consumption at this confidence level.

**Looking at the research question**

- We hypothesized that the insul variable was negatively correlated to the gas consumption this is confirmed by our model, with a coefficient of -1.56. Moreover, it appears to be the most influential covariate, based on the magnitude and significance of their coefficients.

###23.6 Add the estimated parallel lines to scatterplot of the data and comment on the figure. Write the equations that characterize this model.

```{r interaction-plot, eval=FALSE}
# scatterplot with separate fitted lines per group
plot(whiteside$Temp, whiteside$Gas,
     col = as.numeric(Whiteside$Insul),
     pch = 16,
     main = "Interaction visualization",
     xlab = "x1", ylab = "y")

# add fitted lines for each group
for (g in levels(Whiteside$Insul)) {
  subset_g <- whiteside[Whiteside$Insul == g, ]
  abline(lm(y ~ x1, data = subset_g),
         col = which(levels(Whiteside$Insul) == g))
}
legend("topleft", legend = levels(Whiteside$Insul),
       col = 1:nlevels(Whiteside$Insul), lty = 1, cex = 0.8)
```
