---
title: "fac2"
author: "me"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

```{r libraries, eval=FALSE}
# Caricamento librerie usate nel corso
library(MASS)       # stepAIC, Boston dataset
library(car)        # vif, diagnostica
library(corrplot)   # visualizzazione matrici di correlazione
library(boot)       # bootstrap
library(ppcor)      # correlazioni parziali
library(GGally) 
library(mvtnorm)
library(ggplot2)
library(skimr)
library(gridExtra)
```

Data available in water.Rdata11 report concentrations of organic substances in sampled waters. The following records are available: Organic: concentration of organic substances in water (expressed in mg/L); Material: type of material of the
tank storing the water: 0 if plastic, 1 if steel Flow: water flow rate (in liters per minute); Temperature: external temperature (in degrees Celsius).

```{r}
head(water)
```

### 1 Report the sample size and specify if data are derived from an observational study. Summarize the data by the categorical variable and comment on the reported results

```{r}
nrow(water[water$Material==0,])
nrow(water[water$Material==1,])

nrow(water)
```

```{r}
by(water[, c("Organic", "Flow", "Temperature")], water$Material, skim_without_charts)
```
The sample is composed by 388 units, with 323 units where made up of plastic material and 65 of steel.
Comparing the two groups, we can see that the mean of organic in group where the material is plastic is 5.664 versus 7.035 in group 1, we can say that in average there is more organic concentration of substances in steel tanks. Regarding the distributions, we can see that the mean and median are pretty close in both cases, suggesting an approximate symmetrical distribution around the center. 

Regarding the flow variable, we can see that there is a quite big variability in the data: with a standard deviation of 18.634 in the group of plastic material, and 20.272 in the group of steel material. Moreover, we can see that in both cases the median is quite lower than the mean, this suggest a right-skewned distribution, confirmed by the fact that there are some cases where the flow rate is much higher than the median value (max0 = 96 and max1=73). 

Regarding the temperature, we can see that on average the external temperature in the cases of plastic tanks is 13.255 while in the case of steel tanks is 18.172031, higher by around 5 units, and presents also an higher variability, represented by the standard deviation 7.856 vs 5.546721.


### 2 Estimate a multiple linear regression model to explain the concentration of organic substances (Organic) as a function of all the other explanatory variables. Print the results using the summary function and comment on the residuals, the estimated regression coefficients and their standard errors and significance

```{r}
mod_full <- lm(Organic ~ ., data = water)
summary(mod_full)
```
The estimated model is:

$\hat{Y} = 4.058621 + 0.729081 \cdot \text{Material1} +  0.023623 \cdot \text{Flow} + 0.094274 \cdot \text{Temperature}$

**Residual analysis**

The residuals range from $-1.56490$ to $3.14652$ with median $0.04958$, very close to 0, as expected for a well-specified model. The IQR is $[-0.66373, 0.53189]$ approximately centered on 0. However, the maximum 
residual ($3.147$) is substantially larger in absolute value than the minimum ($-1.565$), suggesting a **slight positive asymmetry in the tails**: the model appears to underestimate the probability of `Organic` for one or more units in the sample more than it overestimates it A formal assessment of the model assumptions will be performed in the diagnostic plots section.

**Coefficient interpretation**

- **Material1** ($\hat{\beta}_1 = 0.729081$): significant at the 99.999% level (p-value $2.61e-09$). This means that in the case of the steel tanks we have an average increase of $0.729081$ units in $organic$ with respect to plastic tanks.

- **Flow** ($\hat{\beta}_2 = 0.023623$):  significant at the 0.001% level (p-value < $2.61e-09$). A unit increase is associated with an average increase of $0.023623$ units in $organic$, since the flow is usually a large number, this is very significative.

- **Temperature** ($\hat{\beta}_3 = 0.094274$):  significant at the 0.001% level (p-value < $2.61e-09$). A unit increase is associated with an average increase of $0.094274 $ units in $organic$ concentration.

### 3 Depict the two scatterplots of the residuals with respect to the continuous covariates. Comment on the plots.

```{r}
res <- residuals(mod_full)        # deviance residuals (default for glm)

par(mfrow = c(1, 2))

plot(water$Flow, res,
     main = "Residuals vs Flow",
     xlab = "Flow",
     ylab = "Deviance residuals",
     pch = 16, col = "blue")
abline(h = 0, lty = 2, col = "gray")

plot(water$Temperature, res,
     main = "Residuals vs Temp",
     xlab = "Temp",
     ylab = "Deviance residuals",
     pch = 16, col = "blue")
abline(h = 0, lty = 2, col = "gray")

par(mfrow = c(1, 1))
```

**Residuals vs `flow`**: The residuals are scattered randomly , with no apparent trend 
across the range of the Flow variable`. The dispersion appears  to decrease with the flow covariate, also due to the fact that we have more observations with low flow value.

**Residuals vs `Temp`**: The residuals against Temp show a similar random pattern. Regarding the dispersion, it appears  to remain almost constant within the range of the Temp covariate.



## Es 2

Consider the data in rivers.Rdata2 recording the lengths (in miles) of the 141
major rivers in North America (as compiled by the US Geological Survey).

### .1 Illustrate the data using the appropriate descriptive statistics. Depict the scatter plot and comment according to the applicative context.

```{r}
load("data/rivers.Rdata")
head(rivers)
```
```{r}
skimr::skim_without_charts(rivers)
IQR(rivers)
```

The data set contains 141 observations regarding river lengths, it does not present missing values with a 100% compete rate.
The descriptive statistics for the lenghts are as follows. The mean is 
591.184 and the median is 425: their difference suggests 
that the distribution is quite asymmetric around the central value. In fact, we can say that the distribution is right-skewned, with heavier tails on the higher side, as there are some high values that inflate the mean (ex: maximum value = 3710 miles)

Regarding variability, the standard deviation is 493.87, indicating that the 
observations deviate on average by approximately 493.87 units from the mean, which is a high value and suggest a big variability in the data. 
The interquartile range is 370, with one-quarter of the observations below 
310 and another quarter above 680. The total range is from 135 to 3710, 
approximately 9.662 times the IQR, suggesting that there are extreme values in the tails.

```{r}
plot(rivers, main = "River length distribution")
```
The scatter plot shows that most observations are concentrated between 130 and 680 miles, with few observations over that value that lead to the heavy tail that we have discussed. It confirmed that the distribution appears slightly right-skewed, as a few larger values are present. No strong clustering structure is evident, although there are some isolated observations that may be considered potential outliers. 

### .2 Consider the coefficient of variation (defined as σ/µ) as a parameter of interest. Provide an estimate of its standard error applying nonparametric bootstrap. Comment on the procedure and report the estimated value.

```{r}
library(bootstrap)

set.seed(7369)
n <- length(rivers) 

# funzione ausiliaria: prende indici, restituisce la statistica
aux_fun <- function(ind) {
  sd(rivers[ind])/(mean(rivers[ind]))                      
}

result_boot <- bootstrap::bootstrap(x = 1:n,
                                    nboot = 1000,
                                    theta = aux_fun)

# accesso alle B stime bootstrap
theta_boot <- result_boot$thetastar
summary(theta_boot)

theta_orig = sd(rivers)/mean(rivers)
sd(theta_boot)
```
via the nonparametric bootstrap with $B = 1000$ replications. This procedure 
allows estimation of the standard error and the construction of confidence 
intervals without relying on distributional assumptions, which is particularly The accuracy of the estimator $\hat\theta = (\frac{sigma}{mu})$.

The $B = 1000$ bootstrap estimates of $\hat\theta$ range from 0.537 to 1.074, 
with mean 0.8177 and median 0.8177. Both the mean and median 
of the bootstrap distribution are very close to the original sample estimate 
$\hat\theta = 0.835$, in agreement with the general property that the 
bootstrap distribution is centered (in expectation) on the original sample 
estimate.

The bootstrap standard error is computed as the standard deviation of the $1000$ bootstrap estimates: $\widehat{SE}_{\text{boot}} =  0.0928$. This represents approximately 11,114% of the magnitude of the original 
estimate, suggesting a high level of precision. 
The variability of the bootstrap distribution is also influenced by the presence of few outlier values, which is included in different 
proportions in the various resamples and amplifies the spread of $\hat\theta^*$.


### .3 Provide a bootstrap confidence interval for the parameter at 99%. Depict the histogram of the bootstrap distribution adding the bars referring to the upper and lower bounds of the confidence interval and the mean of the bootstrap replicates along with the value of the estimate on the original sample. Add the legend and comment.

```{r}
ci_99 <- quantile(theta_boot, probs = c(0.005, 0.995))
round(ci_99, 3)
```
```{r}
hist(theta_boot,
     breaks = 40,
     freq = FALSE,                           # densità invece di frequenze
     col = "lightgray", border = "white",
     main = "Bootstrap distribution",
     xlab = "Stima bootstrap",
     ylab = "Densità")

# linea verticale alla stima originale
abline(v = theta_orig, col = "red", lwd = 2, lty = 1)

# linea verticale alla media delle B stime bootstrap
abline(v = mean(theta_boot), col = "blue", lwd = 2, lty = 2)

# se hai calcolato il CI: linee tratteggiate ai bound
ci <- quantile(theta_boot, c(0.025, 0.975))
abline(v = ci, col = "green", lwd = 2, lty = 2)

legend("topright",
       legend = c("Stima originale", "Media bootstrap", "CI bounds"),
       col = c("red", "blue", "green"),
       lty = c(1, 2, 2), lwd = 2,
       cex = 0.7)
```

The histogram of the bootstrap distribution is reported below, with vertical lines indicating the original sample estimate ($[theta orig]$, in red), the mean of the $1000$ bootstrap estimates ($[meanboot]$, in blue), and the lower and upper bounds of the $99$% percentile confidence interval ($0.589$ and $1.031$, in green, dashed).

The shape of the distribution is approximately symmetric and bell-shaped, with the original estimate falling near the center of the bootstrap distribution. 
The confidence interval [$0.589$, $1.031$] covers approximately the central $99$% of the bootstrap mass and represents a range of values centered around the point estimate of the parameter.

It is worth noting that $99%$ is not the probability that the point 
estimate falls within the interval. Rather, we are confident at the 
$99%$ level that the true parameter falls within this range. The bootstrap distribution appears reasonably concentrated around the point estimate, suggesting a fairly precise estimator. In particular, the moderate width of the confidence interval indicates limited sampling variability and therefore a relatively good level of estimation accuracy. No strong asymmetry or multimodality is observed in the bootstrap distribution, supporting the stability of the estimator.


### .4 Illustrate the assumptions under the classical multiple linear regression model.

1. **Linearity**: response is a linear and additive function of covariates
2. **Zero-mean errors**: $\mathbb{E}[\epsilon_i] = 0$
3. **Homoscedasticity**: $\text{Var}(\epsilon_i) = \sigma^2$ constant for all $i$
4. **Independence**: $\text{Cov}(\epsilon_i, \epsilon_j) = 0$ for $i \neq j$
5. **Normality** (needed for t-test, F-test, CI): $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$
