---
title: "fac2"
author: "me"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

```{r libraries, eval=FALSE}
# Caricamento librerie usate nel corso
library(MASS)       # stepAIC, Boston dataset
library(car)        # vif, diagnostica
library(corrplot)   # visualizzazione matrici di correlazione
library(boot)       # bootstrap
library(ppcor)      # correlazioni parziali
library(GGally) 
library(mvtnorm)
library(ggplot2)
library(skimr)
library(gridExtra)
```

Data available in water.Rdata11 report concentrations of organic substances in sampled waters. The following records are available: Organic: concentration of organic substances in water (expressed in mg/L); Material: type of material of the
tank storing the water: 0 if plastic, 1 if steel Flow: water flow rate (in liters per minute); Temperature: external temperature (in degrees Celsius).

```{r}
water = load('data/water.Rdata')
```



